---
title: "Bivariate nodes: joint estimation, residual correlation, and moderation (drm_pair)"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Bivariate nodes: joint estimation, residual correlation, and moderation (drm_pair)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Model-fitting chunks use the drmTMB engine and are eval-gated; the pure-R
# declaration grammar (drm_pair, rho12, corpairs) runs.
has_engine <- requireNamespace("drmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
library(drmSEM)
set.seed(20260828)
```

In standard piecewise structural equation models [@Shipley2000; @Shipley2009;
@Lefcheck2016], each response variable is modelled in its own univariate node.
However, in behavioural ecology, evolutionary genetics, and physiology,
researchers frequently encounter pairs of traits—such as **activity and
boldness** in animal personality [@Brooks2017] or **growth and reproduction** in
life-history tradeoffs—that remain coupled even after conditioning on external
covariates.

This coupling can reflect either **within-observation residual covariance**
($\varepsilon_1 \leftrightarrow \varepsilon_2$) or **between-individual random
covariance** ($u_{1} \leftrightarrow u_{2}$).

`drmSEM` provides the bivariate node constructor **`drm_pair()`** to model these
joint response distributions. This vignette explains:

1. The crucial distinction between **directed paths**, **residual correlations ($\rho_{12}$)**, and **higher-level correlations (`corpairs`)**.
2. Declaring joint bivariate nodes with `drm_pair()`.
3. **Moderation paths on the $\tanh$ link** ($x \to \rho_{12}$, declared as `rho12 = ~ x`).
4. **Basis-set suppression**: why estimating $\rho_{12}$ removes the independence claim from d-separation.
5. Post-estimation inspection: `rho12()`, `corpairs()`, `paths()`, and plotting.

---

## The three ways traits couple

When two traits $Y_1$ and $Y_2$ covary in an observational dataset under an
environmental predictor $X$, there are three fundamentally different mechanisms:

```{r coupling-table, echo = FALSE, results = "asis"}
cat("
| Coupling class | Causal interpretation | Model representation |
| :--- | :--- | :--- |
| **1. Directed path** ($Y_1 \\to Y_2$) | Trait 1 *causes* Trait 2 (e.g. Activity directly induces Boldness). | Univariate piecewise nodes: `boldness ~ activity + x`. |
| **2. Residual correlation** ($\\rho_{12}$) | Unmodelled common shocks or physiology couple $Y_1$ and $Y_2$ at the *observation* level ($\\varepsilon_{1} \\leftrightarrow \\varepsilon_{2}$). | Joint bivariate node via `drm_pair(..., rho12 = ...)`. |
| **3. Between-group correlation** (`corpair`) | Individual syndromes or genetic linkages couple $Y_1$ and $Y_2$ at the *group/individual* level ($u_{\\text{id}, 1} \\leftrightarrow u_{\\text{id}, 2}$). | Shared random intercepts: `(1 \\| id)` in both formulas. |
")
```

Conflating these three processes is a common pitfall. In `drmSEM`, they are
modelled and reported **strictly separately**.

---

## 1. Declaring a bivariate node: `drm_pair()`

`drm_pair()` specifies a joint two-response node. It takes two response formulas,
an optional model for the residual correlation parameter $\rho_{12}$, and optional
grouping levels:

```{r declare-pair}
# Declare a bivariate node for activity and boldness
pair_spec <- drm_pair(
  activity ~ temp + (1 | id),
  boldness ~ temp + (1 | id),
  rho12 = ~ temp    # environmental moderation of coupling strength
)
print(pair_spec)
```

### Accessing declarations before fitting

Before model fitting, `rho12()` and `corpairs()` report the declared structure.
Estimates are honestly reported as `NA` (never fabricated):

```{r pre-fit-accessors}
# Residual within-observation correlation declaration:
rho12(pair_spec)

# Higher-level between-individual random-effect correlation declaration:
corpairs(pair_spec)
```

---

## 2. Moderation paths on the $\tanh$ link ($x \to \rho_{12}$)

In classical SEM and multivariate regression, residual correlation $\rho_{12}$ is
assumed to be constant across all observations. In biological systems, however,
environmental stress or temperature often **modulates the coupling strength**
between traits—for example, coupling may tighten under high temperature and weaken
under benign conditions.

`drmSEM` models $\rho_{12}$ as a first-class distributional component on the
Fisher $z$ ($\text{atanh}$) link:

$$\text{atanh}(\rho_{12}) = \beta_0 + \beta_1 x$$

$$\rho_{12}(x) = \tanh(\beta_0 + \beta_1 x) = \frac{\exp(2(\beta_0 + \beta_1 x)) - 1}{\exp(2(\beta_0 + \beta_1 x)) + 1} \in (-1, 1)$$

When declared via `rho12 = ~ x`, this edge represents a **directed path into the
residual correlation component**. It appears in `paths(sem)` under `component = "rho12"`,
distinguishing it from both mean effects and $Y_1 \to Y_2$ regressions.

---

## 3. Joint fitting with `drm_sem()`

When passed to `drm_sem()`, a `drm_pair()` declaration is fitted as a single
joint bivariate model in the `drmTMB` engine:

```{r fit-bivariate, eval = has_engine}
# Simulate bivariate location-scale-correlation data
n <- 400
temp <- rnorm(n)
# Temperature moderates residual correlation via tanh link:
# eta_rho = 0.20 + 0.50 * temp
eta_rho <- 0.20 + 0.50 * temp
rho_true <- 0.999 * tanh(eta_rho)

# Generate coupled residuals
e1 <- rnorm(n)
e2 <- rho_true * e1 + sqrt(pmax(1 - rho_true^2, 1e-6)) * rnorm(n)

dat <- data.frame(
  temp = temp,
  activity = 0.5 * temp + e1,
  boldness = 0.3 * temp + e2
)

# Downstream fitness response depending on both traits
dat$fitness <- 0.6 * dat$activity + 0.4 * dat$boldness + rnorm(n, sd = 0.5)

# Fit the SEM with joint bivariate node + downstream fitness node:
sem_biv <- drm_sem(
  pair = drm_pair(
    activity ~ temp,
    boldness ~ temp,
    rho12 = ~ temp
  ),
  fitness = drm_node(
    drmTMB::bf(fitness ~ activity + boldness),
    family = stats::gaussian()
  ),
  data = dat
)

summary(sem_biv)
```

---

## 4. Post-estimation accessors

Once fitted, `rho12()` and `paths()` report the estimated correlation coefficients
with standard errors, $z$-statistics, and $p$-values on the $\tanh$ link scale:

```{r fitted-accessors, eval = has_engine}
# Residual correlation parameter estimates on tanh link
rho12(sem_biv)

# Full structural paths table (includes directed paths to mu and to rho12)
paths(sem_biv)
```

---

## 5. Basis-set suppression in d-separation

In an ordinary piecewise SEM without a bivariate node, omitting an arrow between
`activity` and `boldness` asserts that they are conditionally independent given
`temp` ($Y_1 \perp\mkern-10mu\perp Y_2 \mid X$). This claim would enter the
d-separation basis set and be tested via Fisher's C [@Shipley2000].

When `activity` and `boldness` are declared as a `drm_pair()`, **the residual
covariance $\rho_{12}$ is explicitly estimated**. Consequently:

- The conditional independence claim `activity _||_ boldness | temp` is **honestly suppressed** from `basis_set(sem)`.
- Fisher's C test tests only genuine conditional independencies outside the joint pair.

```{r basis-set-check, eval = has_engine}
basis_set(sem_biv)
```

Notice that `activity` versus `boldness` does not appear in the basis set.

---

## Summary of best practices

1. **Distinguish the three couplings**: Use a directed arrow $Y_1 \to Y_2$ for causal causation; use `drm_pair()` for residual correlation ($\rho_{12}$) and shared random effects (`corpairs`).
2. **Model moderated coupling**: Use `rho12 = ~ x` when environmental factors alter the tightness of correlation between traits.
3. **Interpret link scales**: Remember that $\rho_{12}$ regression coefficients operate on the $\text{atanh}$ link scale; transform back via $\tanh(\cdot)$ to obtain correlations on $[-1, 1]$.
4. **Inspect d-separation**: Verify with `basis_set(sem)` that estimated covariance pairs are properly excluded from conditional independence tests.

---

## References
