---
title: "Bivariate nodes: two responses, one correlation (drm_pair)"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Bivariate nodes: two responses, one correlation (drm_pair)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# drmSEM never fits its own likelihoods. The bivariate JOINT fit that would
# estimate rho12 inside one drmTMB model is the 0.4 engine deliverable, and the
# dev lane has no engine. So every chunk that would need a live drmTMB fit is
# shown for illustration only (`eval = has_engine`, FALSE here). The pure-R
# declaration grammar — drm_pair(), rho12(), corpairs(), drm_expand_pair() — runs
# normally, and those chunks are marked `eval = TRUE` explicitly.
has_engine <- FALSE
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_engine)
library(drmSEM)
```

If you fit GLMMs [@Brooks2017], you already know how to ask "does `x` change
the *mean* of `activity`?" This vignette is about a different, sharper question
that a mean-only model cannot phrase: **do two responses stay coupled after we
condition on their predictors, and is that coupling causal or merely
correlational?** The grammar follows the bidirected-edge convention of
@Bollen1989 and @Shipley2016 and is exposed in the bivariate-node helper
`drm_pair()`.

It is the companion to the
[covariance-edges vignette](covariance-edges-and-composites.html), which
introduces the `covary()` / `covariances()` primitive. Here we work the *node*
constructor — `drm_pair()` — through a concrete animal-personality example, and
we are scrupulous about one thing throughout: **the correlation estimates are
`NA` by construction.** Declaring a bivariate node is not fitting one. The
design of record is `docs/design/07-bivariate-covariance-edges.md`.

## A directed path is not a covariance edge

Two behavioural traits — **activity** (how much an animal moves) and
**boldness** (how readily it explores a novel object) — are measured repeatedly
on the same individuals (grouping `id`), under some environmental predictor `x`
(say, temperature). There are three genuinely different things you might mean by
"activity and boldness go together", and conflating them is exactly the mistake
this grammar exists to prevent.

1. **A directed path `activity -> boldness`.** You believe moving more *causes*
   an animal to act bolder. This is an ordinary piecewise path: `boldness` is one
   node, `activity` is a predictor in it, and `paths()` / `dsep()` / the effect
   functions already handle it. **No bivariate node is needed.** Rule of thumb: if
   you can name a direction, it is a directed path.

2. **A residual correlation `rho12` (within-observation).** You make *no*
   directional claim. You only mean: on a single occasion, when this individual's
   activity is above its own predicted mean, is its boldness *also* above its
   predicted mean? This is the within-observation coupling `eps_activity <->
   eps_boldness` that remains after each trait's mean and scale are accounted for.
   It is a double-headed arc, not an arrow.

3. **A higher-level correlation `corpair` (between-individual).** Again no
   direction — but now at the level of the *individual*, not the occasion: do
   individuals that *average* high on activity also *average* high on boldness?
   This is the random-effect coupling `u_id,activity <-> u_id,boldness` among the
   `(1 | id)` intercepts the two responses share.

Classes 2 and 3 are not interchangeable, and an animal-personality study is the
classic case where they diverge: bold individuals can be more active *on average*
(a positive `corpair`) while, occasion by occasion, the residual `rho12` is near
zero — or the reverse. `drmSEM` reports them **separately** and never collapses
one into the other.

> The same separation drives the *behaviour* of the package: a directed
> `x -> rho12` path is a legitimate edge into the correlation component and flows
> through `paths()` / effects; the `rho12` and `corpair` *covariance* arcs carry
> no direction and no mediated effect, but they do drop the
> `activity _||_ boldness` independence claim from the d-separation basis set.

## Declaring a bivariate node with `drm_pair()`

`drm_pair()` is the bivariate counterpart of `drm_node()`: two response formulas,
two families, an optional model for the residual correlation itself, and an
auto-detected higher-level edge wherever the two formulas share a grouping
factor. It is **pure R** and runs without any engine, so the chunks in this
section evaluate normally.

```{r pair-decl, eval = TRUE}
pair <- drm_pair(
  activity ~ x + (1 | id),
  boldness ~ x + (1 | id),
  rho12 = ~ x                 # a directed path INTO the correlation component
)
```

Note what `rho12 = ~ x` means. It is **not** `activity -> boldness`, and it is
**not** a mean effect on either trait. It says: the *residual correlation*
between activity and boldness is itself allowed to change with `x` — perhaps the
two traits couple more tightly at high temperature. That is a directed path into
the `rho12` component, on its `tanh` link scale. Leave `rho12` as the default
`NULL` for a constant residual correlation.

Printing the pair shows the whole declaration at a glance, and — prominently —
the honest status line:

```{r pair-print, eval = TRUE}
print(pair)
```

The two responses each share a `(1 | id)` grouping, so `drm_pair()`
**auto-detects** a higher-level `corpair` edge at the `id` level. You can force a
specific level with `level = "id"`, or suppress the higher-level edge entirely
(residual `rho12` only) with `level = NA`.

## Reading the declared edges: `rho12()` and `corpairs()`

Two accessors report the declared structure, each kept strictly separate from
`paths()`. `rho12()` returns the residual (within-observation) edge:

```{r rho12, eval = TRUE}
rho12(pair)
```

`corpairs()` returns the higher-level (between-individual) edge(s):

```{r corpairs, eval = TRUE}
corpairs(pair)
```

Look at the `estimate` column in both: it is **`NA`**. This is the core honesty
of the grammar, and it is deliberate, not a placeholder waiting to be silently
filled.

> **`drm_pair()` never fabricates a correlation.** drmSEM is piecewise and does
> not fit its own likelihoods. Estimating `rho12` (and any `corpair`) means
> fitting *both responses jointly* inside **one** bivariate `drmTMB` model and
> reading the fitted correlation back — that joint fit is the **0.4 engine
> deliverable**. Until a live bivariate fit is supplied, the declared structure
> is complete but every estimate is `NA` by construction. The accessors carry a
> grey note saying exactly this. Anyone who shows you a number here from the
> grammar layer alone is showing you something `drmSEM` did not compute.

This is the same boundary as in the covariance-edges vignette, made
node-shaped: the *grammar and graph semantics* ship today; the *fitted estimate*
is the engine step.

## Expanding a pair onto the shipped grammar: `drm_expand_pair()`

How does a bivariate declaration connect to the rest of `drmSEM`?
`drm_expand_pair()` bridges it onto the pieces that already ship: two ordinary
`drm_node()` sub-nodes (one per response) plus the `covary()` covariance edges
(the residual `rho12` and any higher-level `corpair`).

Building the `drm_node()` objects wraps each formula with `drmTMB::bf()`, so this
one helper needs `drmTMB` available even though it does not *fit*. We gate it on
`requireNamespace()` and skip cleanly when the engine is absent:

```{r expand, eval = requireNamespace("drmTMB", quietly = TRUE)}
expanded <- drm_expand_pair(pair)
names(expanded$nodes)     # "activity" "boldness" — two ordinary sub-nodes
expanded$covariances      # the residual rho12 edge + the id-level corpair edge
```

This is the **documented hook point for the 0.4 engine lane**: when the joint
bivariate fit lands, it swaps the two independent node fits for one joint fit,
while the covariance edges and the `rho12()` / `corpairs()` accessors stay
exactly the same — they simply start returning a non-`NA` estimate read back from
the fit.

## Putting the pair in a SEM, and seeing the arcs

In a full model the expanded pair sits alongside ordinary nodes. Suppose
downstream **fitness** depends on both behavioural traits. Assembling and *fitting*
that SEM needs the engine, so the chunk below is illustrative only:

```{r sem, eval = has_engine}
sem <- drm_sem(
  activity = drm_node(drmTMB::bf(activity ~ x + (1 | id)),
                      family = stats::gaussian()),
  boldness = drm_node(drmTMB::bf(boldness ~ x + (1 | id)),
                      family = stats::gaussian()),
  fitness  = drm_node(drmTMB::bf(fitness ~ activity + boldness),
                      family = stats::gaussian()),
  data = dat,
  covariances = covary("activity", "boldness")        # residual rho12 edge
)

paths(sem)         # directed-only: x -> activity, x -> boldness, activity -> fitness, ...
covariances(sem)   # the covariance edges, reported SEPARATELY from paths()
rho12(sem)         # the declared residual edge (estimate still NA without a joint fit)
```

`plot(sem, show = "all")` draws the three edge classes so they are visually
distinct, mirroring `docs/design/07-bivariate-covariance-edges.md`:

```{r plot, eval = has_engine, fig.alt = "DAG of activity, boldness and fitness with directed arrows and a double-headed residual covariance arc between activity and boldness."}
plot(sem, show = "all")
```

Conceptually:

- **Solid coloured arrows** are directed paths, styled by the distributional
  component they target (`mu` black, `sigma` green, ..., and a directed
  `x -> rho12` path as a long dash).
- A **solid grey double-headed arc** is a residual `rho12` covariance edge —
  `eps_activity <-> eps_boldness`, no direction.
- A **dashed grey double-headed arc** is a higher-level `corpair` edge —
  `u_id,activity <-> u_id,boldness`, again no direction.

`show = "paths"` suppresses the covariance arcs and draws the directed structural
edges only, when you want the bare causal skeleton.

## Recap

- A **directed path** `activity -> boldness` makes a causal claim and is an
  ordinary node + predictor — no bivariate node needed.
- A **bivariate node** (`drm_pair()`) declares a *correlation* between two
  responses with **no direction**: a residual `rho12` (within-observation,
  `eps <-> eps`) and, where the responses share a grouping, a higher-level
  `corpair` (between-unit, `u <-> u`). The two answer different biological
  questions and are reported separately by `rho12()` and `corpairs()`.
- `rho12 = ~ x` is a *directed path into the correlation component*, distinct from
  both a mean effect and a `y1 -> y2` arrow.
- **The estimates are `NA` by construction.** The grammar declares the structure;
  the fitted correlation must be read back from one joint bivariate `drmTMB` fit
  (the 0.4 engine step). `drm_pair()` never fabricates a number.
- `drm_expand_pair()` bridges the declaration onto the shipped `covary()` grammar
  and is the hook point where the joint fit will land.

For the covariance-edge primitive (`covary()` / `covariances()`) and how a
declared edge changes d-separation, see
[Covariance edges, composites, and path attribution](covariance-edges-and-composites.html).
For the full design — the three edge classes, residual vs higher-level, the
level-compatibility rule, and the honest engine boundary — see
`docs/design/07-bivariate-covariance-edges.md`.

## References
