---
title: "Calibration of the d-separation test (OQ-6)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibration of the d-separation test (OQ-6)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# This is a PRECOMPUTED vignette. Every result shown below is read from a
# cached study object built in the live-drmTMB lane (see the "How this was
# generated" section). The global eval = FALSE guarantees that R CMD check,
# pkgdown, and any reader rebuilding the vignette NEVER fit a model here.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)

source_results_path <- file.path("inst", "calibration",
                                 "calibration-results.rds")
installed_results_path <- system.file("calibration", "calibration-results.rds",
                                      package = "drmSEM")
results_path <- if (file.exists(source_results_path)) {
  source_results_path
} else {
  installed_results_path
}
cal <- if (nzchar(results_path) && file.exists(results_path)) {
  readRDS(results_path)
} else {
  NULL
}
```

All results below are **precomputed on live drmTMB; this document performs no
model fitting.** The study object is regenerated only in the drmTMB lane and
ships as a cached `.rds` inside the package.

```{r guard, eval = TRUE, echo = FALSE, results = "asis"}
# Graceful degrade: render an explanatory note (not an error) and stop early if
# the cache has not been built yet, so the doc still renders in the
# no-engine lane.
if (is.null(cal)) {
  cat(
    "> **The calibration study has not yet been run in the live-drmTMB lane.**\n",
    ">\n",
    "> This page is a precomputed scaffold. Once",
    "`inst/calibration/generate.R`",
    "is run with the drmTMB engine, the cached results will appear here:",
    "Type-I tables, the stratified Type-I figure, Fisher's C uniformity, and",
    "power curves. The study design is shown below for transparency.\n",
    sep = " "
  )
  knitr::knit_exit()
}
```

## What this page certifies

The d-separation machinery ([`dsep()`], [`fisher_c()`]) tests each missing
arrow on **every modelled distributional component** of the response. Whether
that any-component test is *calibrated* -- nominal Type-I error under a
correctly specified DAG, and adequate power against a real omitted edge -- is an
empirical question. This page is the **recovery / coverage-tier** evidence for
that question.

It is deliberately distinct from the fast smoke check in
`tests/testthat/test-calibration.R` (V-17), which runs only 20 reps of a single
chain to catch gross regressions in CI. That smoke test is **not** calibration
evidence; this study is. Until the cache here meets all five acceptance criteria
below, the d-separation test remains labelled **experimental**.

## The study design (shown, never run)

The grid sweeps four sample sizes, six omitted-edge strengths, and three
data-generating-process (DGP) families, with many replicates per cell.

```{r design-grid, echo = TRUE, eval = FALSE}
GRID <- list(
  n      = c(100, 250, 500, 1000),
  beta   = c(0, 0.1, 0.2, 0.3, 0.5, 0.8),   # strength of the OMITTED x -> y edge
  family = c("mean_only", "distributional", "cross_link"),
  reps   = 200
)
```

Each DGP is a chain `x -> m -> y` whose true graph also contains a direct
`x -> y` edge of strength `beta`. We fit the SEM **without** that edge, so the
basis-set claim `x _||_ y | {m}` is true exactly when `beta == 0` (Type-I) and
false when `beta > 0` (power). The three families differ in *which
distributional component* the omitted edge enters:

```{r design-dgp, echo = TRUE, eval = FALSE}
# mean-only: omitted edge perturbs the MEAN of y
y <- rnorm(n, beta * x + 0.7 * m, 1)

# distributional: omitted edge perturbs the SPREAD (sigma) of y, not its mean
y <- rnorm(n, 0.7 * m, exp(beta * x))   # could equally drive zi for a count y

# cross-link: m is distributionally driven by x (its sigma), omitted edge on
# the mean of y -- exercises an augmented refit over a multi-component node
m <- rnorm(n, 0.5 * x, exp(0.3 * x))
y <- rnorm(n, beta * x + 0.7 * m, 1)
```

```{r design-loop, echo = TRUE, eval = FALSE}
# For each cell and replicate: build the SEM, run dsep(), and record the
# x _||_ y | {m} claim's p-value, its df (= number of augmented components),
# its status, and the per-replicate Fisher's C p-value. (Schematic; the
# authoritative version is inst/calibration/generate.R.)
sem <- drm_sem(
  m = drm_node(drmTMB::bf(m ~ x), family = gaussian()),
  y = drm_node(drmTMB::bf(y ~ m), family = gaussian()),
  data = dat
)
d  <- dsep(sem)
p  <- d$p.value[d$x == "x" & d$y == "y"]
df <- d$df[d$x == "x" & d$y == "y"]          # augmented-component count
fc <- attr(d, "fisher_c")$p.value
```

This study code is shown **for transparency only** and is never evaluated when
the vignette is built.

## Acceptance criteria for promoting V-17

The cached study object stores these checks in `cal$acceptance`. V-17 can move
from experimental to validated only if all five pass:

1. Every family x n x beta cell has at least 95% usable d-separation claims
   (`status == "ok"` and finite p-value).
2. Every beta = 0 family x n Type-I estimate lies inside the 99% binomial
   Monte-Carlo band around the nominal alpha.
3. Every beta = 0 family x augmented-component-count (`claim_df`) Type-I
   estimate lies inside the same 99% Monte-Carlo band.
4. Fisher's C p-values under the null are close to Uniform(0, 1): one-sample
   KS p >= 0.01 and median p in [0.40, 0.60].
5. Power behaves as expected: beta = 0.8 gives power >= 0.80 in every family x n
   cell; beta = 0.5 gives power >= 0.70 for n >= 250; and power is nondecreasing
   aside from at most 0.05 Monte-Carlo jitter.

```{r acceptance-table, echo = FALSE, eval = !is.null(cal)}
knitr::kable(
  cal$acceptance$criteria,
  col.names = c("Criterion", "Rule", "Pass"),
  caption = "Acceptance checks for promoting V-17."
)
```

## Type-I error by family and sample size

```{r type1-table, echo = FALSE, eval = !is.null(cal)}
type1_tab <- cal$type1[order(cal$type1$family, cal$type1$n),
                       c("family", "n", "reject", "mc_lower", "mc_upper",
                         "n_rep", "pass")]
knitr::kable(
  type1_tab,
  digits  = 3,
  col.names = c("DGP family", "n", "Type-I", "MC lower", "MC upper",
                "reps", "pass"),
  caption = sprintf(
    "Empirical Type-I rate at nominal alpha = %.2f, with the 99%% Monte-Carlo acceptance band.",
    cal$meta$alpha
  )
)
```

### The centerpiece: Type-I stratified by augmented-component count

The any-component refit adds the candidate predictor to **every** modelled
component of the response, so the claim's degrees of freedom equal the number of
augmented components. Mis-calibration, if it exists, is most likely to show up
as the component count grows -- which is exactly what this figure isolates.

```{r type1-by-df, echo = FALSE, eval = !is.null(cal) && requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 3.6, fig.alt = "Empirical Type-I rate plotted against the number of augmented distributional components, faceted by DGP family, with a reference line at the nominal 0.05 level."}
library(ggplot2)
ggplot(cal$type1_by_df,
       aes(x = factor(claim_df), y = reject)) +
  geom_hline(yintercept = cal$meta$alpha, linetype = 2, colour = "grey50") +
  geom_point(size = 2, colour = "#0f7887") +
  facet_wrap(~ family) +
  labs(
    x = "augmented-component count (claim df)",
    y = "empirical Type-I rate",
    title = "Type-I error stratified by augmented-component count",
    subtitle = sprintf("dashed line = nominal alpha = %.2f", cal$meta$alpha)
  ) +
  theme_minimal()
```

## Fisher's C uniformity under the null

Under a correctly specified DAG (`beta == 0`), the per-replicate Fisher's C
p-value should be Uniform(0, 1). Departures from the diagonal in the QQ plot
flag mis-calibration of the combined test.

```{r fisher-c-qq, echo = FALSE, eval = !is.null(cal) && requireNamespace("ggplot2", quietly = TRUE), fig.width = 5, fig.height = 5, fig.alt = "Quantile-quantile plot of null Fisher's C p-values against the uniform distribution, points lying close to the 45-degree reference line."}
library(ggplot2)
p_null <- sort(cal$fisher_c_null_p)
qq <- data.frame(
  theoretical = stats::ppoints(length(p_null)),
  observed    = p_null
)
ggplot(qq, aes(x = theoretical, y = observed)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(size = 1, colour = "#0f7887", alpha = 0.7) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x = "theoretical uniform quantile",
    y = "observed Fisher's C p-value",
    title = "Fisher's C p-value uniformity under the null"
  ) +
  theme_minimal()
```

## Power curves

Rejection rate against the omitted-edge strength `beta`, faceted by sample size
and DGP family. A well-behaved test climbs from the nominal level at `beta = 0`
toward 1 as the omitted edge strengthens.

```{r power-curves, echo = FALSE, eval = !is.null(cal) && requireNamespace("ggplot2", quietly = TRUE), fig.width = 7.5, fig.height = 5, fig.alt = "Power curves showing rejection rate rising with omitted-edge strength beta, faceted in a grid by sample size and DGP family."}
library(ggplot2)
ggplot(cal$power, aes(x = beta, y = reject, colour = factor(n))) +
  geom_hline(yintercept = cal$meta$alpha, linetype = 2, colour = "grey60") +
  geom_line() +
  geom_point(size = 1.5) +
  facet_wrap(~ family) +
  labs(
    x = "omitted-edge strength (beta)",
    y = "rejection rate (power)",
    colour = "n",
    title = "Power vs omitted-edge strength"
  ) +
  theme_minimal()
```

## How this was generated

The cache backing this page is built by **`inst/calibration/generate.R`**, run
in the live-drmTMB lane:

```sh
Rscript inst/calibration/generate.R
```

That script guards on `requireNamespace("drmTMB")`, sweeps the full grid above,
collects only data (per-claim p-values, df, status, and per-replicate Fisher's C
p-values -- never fitted objects), and writes a compact summary list to
`inst/calibration/calibration-results.rds`. The cache carries a `meta` stamp
recording the **drmTMB version, date, R version, and git SHA** so every figure
on this page is traceable to the exact engine that produced it.

```{r provenance, echo = FALSE, eval = !is.null(cal)}
m <- cal$meta
cat(sprintf(
  "Cache provenance:\n  generated:      %s\n  drmTMB version: %s\n  drmTMB SHA:     %s\n  drmSEM version: %s\n  R version:      %s\n  git SHA:        %s\n  reps per cell:  %s\n",
  m$generated_at, m$drmTMB_version,
  if (is.null(m$drmTMB_remote_sha)) NA_character_ else m$drmTMB_remote_sha,
  m$drmSEM_version, m$R_version, m$git_sha, m$reps
))
```
