---
title: "Covariance edges, composites, and path attribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covariance edges, composites, and path attribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Every model-fitting chunk uses the drmTMB engine (compiled from source), so the
# chunks below are illustrative and not executed; `eval = has_engine` keeps them
# from running. The pure-R declaration helpers (covary(), drm_composite()) DO run.
has_engine <- FALSE
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_engine)
library(drmSEM)
```

This vignette covers three pieces of `drmSEM` that go *beyond* a plain directed
DAG of distributional paths:

1. **Covariance edges** — two responses that stay coupled after their predictors,
   declared with `covary()` and reported by `covariances()`.
2. **Composite constructs** — a weighted-sum or principal-component index of
   observed indicators, declared with `drm_composite()`.
3. **Path attribution** — splitting an indirect effect by mediator and by
   distributional channel with `path_effects()`.

## 1. Covariance edges: `covary()` and `covariances()`

A directed path says "`X` causes a component of `Y`". A **covariance edge** says
something different: two responses are *allowed to remain associated* after their
modelled predictors, with **no direction and no mediated effect**. drmSEM keeps
the two ideas strictly separate (`docs/design/07-bivariate-covariance-edges.md`).

There are two kinds, and they answer different biological questions:

- a **residual** correlation `rho12` (within-observation, `eps_y1 <-> eps_y2`):
  "on a single occasion, when this individual's `activity` is above its predicted
  mean, is its `boldness` also above its predicted mean?";
- a **higher-level** random-effect correlation `corpair` (between-unit,
  `u_id,y1 <-> u_id,y2`): "do individuals that *average* high on `activity` also
  average high on `boldness`?".

`covary()` is the pure-R declaration primitive (it runs without the engine):

```{r covary, eval = TRUE}
# residual (rho12) edge:
covary("activity", "boldness")
# higher-level (corpair) edge sharing the `id` grouping:
covary("activity", "boldness", level = "id")
```

Pass declarations to `drm_sem()` / `drm_psem()` via `covariances =`. They are
reported by `covariances()` — kept **out** of `paths()`, which stays
directed-only — and they make `basis_set()` / `dsep()` drop the
`activity _||_ boldness` independence claim (Shipley's bidirected-edge rule),
because the model explicitly allowed the two to stay coupled.

```{r covariances}
sem <- drm_sem(
  activity = drm_node(drmTMB::bf(activity ~ temp), family = stats::gaussian()),
  boldness = drm_node(drmTMB::bf(boldness ~ temp), family = stats::gaussian()),
  data = dat,
  covariances = covary("activity", "boldness")
)
covariances(sem)   # the residual rho12 edge, separate from paths()
paths(sem)         # directed-only: temp -> activity, temp -> boldness
dsep(sem)          # the activity _||_ boldness claim is dropped
```

> The grammar and the d-separation consequence ship today. A first-class
> bivariate node constructor (`drm_pair()`), `rho12()` / `corpairs()` read-back
> from a live fit, and double-headed-arc plotting need a bivariate `drmTMB` fit
> and are on the roadmap.

## 2. Composite constructs: `drm_composite()` and `loadings()`

A **composite (formative) construct** is a deterministic index built from two or
more observed indicators *before* fitting — a weighted sum (`method = "fixed"`)
or the first principal-component score (`method = "pca"`). Once materialized it is
an ordinary column, so a node formula can use it as a predictor or response, with
**no engine change**. `drm_composite()` runs in pure R:

```{r composite, eval = TRUE}
dat_ind <- data.frame(len = rnorm(50), mass = rnorm(50), wing = rnorm(50))
body_size <- drm_composite("body_size", c("len", "mass", "wing"),
                           method = "pca", data = dat_ind)
body_size
```

Pass composites to `drm_sem()` via `composites =`; the construct column is built
before fitting, and `loadings()` reports the indicator weights (kept separate from
the structural `paths()`):

```{r composite-sem}
sem2 <- drm_sem(
  fitness = drm_node(drmTMB::bf(fitness ~ body_size + temp),
                     family = stats::gaussian()),
  data = dat,
  composites = drm_composite("body_size", c("len", "mass", "wing"), data = dat)
)
loadings(sem2)
```

> This is the *formative* construct (indicators define the index). *Reflective*
> latent variables — a latent common cause with a measurement model — need a joint
> likelihood drmTMB does not fit piecewise, and are out of scope here
> (`docs/design/09-latent-variables.md`).

## 3. Path attribution: `path_effects()`

`indirect_effects()` routes through a *set* of mediators. `path_effects()`
attributes that indirect effect to **each mediator** and to each mediator's
**distributional channel**.

By mediator (`by = "mediator"`, the default), each gets an **inclusion** effect
(only that mediator responds) and an **exclusion** effect (its marginal given the
others), plus an explicit `interaction_remainder` — the pieces sum to the total
only when the mediators act additively, and the remainder is reported rather than
hidden:

```{r path-mediator}
path_effects(sem3, from = "temp", to = "survival")
```

By component (`by = "component"`), each mediator's effect is split into a
`mean_channel` plus one channel per non-mean component (`sigma_channel`,
`zi_channel`, ...) — the drop in the effect when that component is frozen at its
reference value — and a `component_remainder` for the part that does not separate
cleanly (under a nonlinear outcome the channels are not an exact partition, so the
remainder is reported rather than hidden):

```{r path-component}
path_effects(sem3, from = "temp", to = "survival", by = "component")
```

This is a **model-based decomposition**, not a claim of nonparametric
path-specific identification (which requires the recanting-witness criterion). The
cross-world natural per-mediator variant is on the roadmap. See
`docs/design/02-effect-calculus.md`.
