---
title: "Distributional piecewise SEM with drmSEM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distributional piecewise SEM with drmSEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# drmSEM never fits its own likelihoods: every endogenous node is one drmTMB
# fit, and drmTMB compiles TMB (C++) from source. While the drmTMB-integration
# layer is still pending end-to-end validation (see
# docs/memory/VALIDATION_LEDGER.md), the model-fitting chunks below are shown
# for illustration but not evaluated. Flip `has_engine` to
# requireNamespace("drmTMB", quietly = TRUE) once integration is validated.
has_engine <- FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_engine
)
library(drmSEM)
```

## Why a *distributional* piecewise SEM?

Most structural equation modelling treats a path `X -> Y` as a statement about
the **expected value** of `Y`: increase `X`, and the mean of `Y` shifts. That is
only one of the things a predictor can do. A predictor can also change how
*variable* a response is, how *over-dispersed* it is, how often it is a
*structural zero*, or how strongly two responses *co-vary in their residuals*.

`drmSEM` makes those targets first-class. It is the **SEM / graph /
d-separation / path / effect-decomposition layer**; the
[`drmTMB`](https://github.com/itchyshin/drmTMB) package is the **fitting
engine**. Each endogenous node is exactly one `drmTMB` fit, the system is
**piecewise**, and the graph must be a **DAG**. A causal path is
**component-labelled**: it targets one modelled distributional component of a
node:

| Component  | Symbol      | What a path to it means                          |
| ---------- | ----------- | ------------------------------------------------ |
| mean       | `mu`        | shifts the expected response                     |
| scale      | `sigma`     | changes residual scale / dispersion, not the mean|
| shape      | `nu`        | changes a shape parameter (e.g. tail/skew)       |
| zero-infl. | `zi`        | changes the probability of a structural zero     |
| hurdle     | `hu`        | changes the probability of crossing the hurdle   |
| RE scale   | `sd(group)` | changes among-group heterogeneity                |
| residual r | `rho12`     | changes the bivariate residual correlation       |

The cardinal rule, enforced throughout the package: **a path to a non-`mu`
component is never reported as a mean effect.** A `temp -> sigma(size)` path is
a statement about *spread*, and `drmSEM` keeps it labelled as such everywhere --
in `paths()`, in `plot()`, and in the effect decomposition.

## A worked system: `size -> abundance -> survival`

We use a three-node ecological chain. Conceptually:

- **`size`** of organisms depends on temperature and habitat. We also let
  temperature act on the *scale* of size: `sigma(size) ~ temp`. Equation:
  `size ~ Normal(mu = b0 + b1*temp + b2*habitat, sigma = exp(g0 + g1*temp))`.
- **`abundance`** is a count depending on `size` and `temp`, modelled with a
  negative binomial, with habitat acting on *zero-inflation*:
  `zi(abundance) ~ habitat`. Equation (link scale):
  `log(mu) = a0 + a1*size + a2*temp`, `logit(zi) = c0 + c1*habitat`.
- **`survival`** is a binomial proportion (`cbind(alive, dead)`) depending on
  `abundance` and `size`, with a beta-binomial family to absorb extra-binomial
  variation: `logit(mu) = d0 + d1*abundance + d2*size`.

Temperature is exogenous and reaches `survival` only indirectly -- through the
*mean*, the *scale*, and the *zero-inflation* of the intermediate nodes. That is
exactly the situation where coefficient-product mediation breaks down and a
distribution-aware, simulation-based decomposition is required.

### Simulate the data

```{r simulate, eval = TRUE}
# This chunk uses only base R, so it always runs.
set.seed(1)
n <- 400L
habitat <- factor(sample(c("A", "B"), n, replace = TRUE))
temp <- rnorm(n)

# size: mean depends on temp + habitat; SCALE (sigma) depends on temp.
mu_size <- 2 + 0.8 * temp + 0.5 * (habitat == "B")
sd_size <- exp(-0.2 + 0.4 * temp)            # temp -> sigma(size)
size <- rnorm(n, mu_size, sd_size)

# abundance: NB mean depends on size + temp; ZI depends on habitat.
mu_ab <- exp(0.5 + 0.3 * size + 0.2 * temp)
zi_ab <- plogis(-1 + 1.2 * (habitat == "B"))  # habitat -> zi(abundance)
abundance <- ifelse(rbinom(n, 1, zi_ab) == 1, 0,
                    rnbinom(n, mu = mu_ab, size = 2))

# survival: binomial proportion out of a fixed number of trials.
trials <- 20L
p_surv <- plogis(-1 + 0.05 * abundance + 0.15 * size)
alive <- rbinom(n, trials, p_surv)
dead <- trials - alive

dat <- data.frame(temp, habitat, size, abundance, alive, dead)
head(dat)
```

The data-generating process puts a real signal into a *non-mean* component of
two different nodes (`sigma(size)` and `zi(abundance)`), which is what we want
the SEM to recover and propagate.

### Build the SEM

`drm_sem()` takes one named `drm_node()` per endogenous response and a shared
`data` frame. Each `drm_node()` wraps a `drmTMB::bf()` formula (which can carry
component sub-formulas like `sigma ~ temp` or `zi ~ habitat`) and a family.

```{r build}
sem <- drm_sem(
  size      = drm_node(drmTMB::bf(size ~ temp + habitat, sigma ~ temp),
                       family = stats::gaussian()),
  abundance = drm_node(drmTMB::bf(abundance ~ size + temp, zi ~ habitat),
                       family = drmTMB::nbinom2()),
  survival  = drm_node(drmTMB::bf(cbind(alive, dead) ~ abundance + size),
                       family = drmTMB::beta_binomial()),
  data = dat
)
```

If you have already fitted nodes yourself, `drm_psem()` assembles a SEM from
existing `drmTMB` objects instead of fitting internally; both routes produce the
same object.

### Inspect the component-labelled paths

`paths()` returns one row per fitted fixed-effect coefficient across all nodes,
each tagged with the **component** it targets and the **link** on which it acts.
This is where the distributional structure becomes legible: the `temp` path into
`size` appears twice -- once on `mu`, once on `sigma` -- and they are not the
same claim.

```{r paths}
paths(sem)
```

`check_sem()` reports, per node, the family, the modelled components,
convergence, whether a fixed-effect covariance is available (needed for
d-separation and effect intervals), and whether a realized-value sampler exists
(needed for fully distribution-mediated effects).

```{r check}
check_sem(sem)
```

### See the DAG

`plot()` draws the graph with **component-labelled edges**, so a `sigma` or `zi`
arrow is visually distinct from a `mu` arrow rather than collapsed into a generic
"effect".

```{r plot, fig.alt = "Directed acyclic graph of the size, abundance and survival nodes with component-labelled edges."}
plot(sem)
```

## d-separation and Fisher's C

A piecewise SEM is tested by checking the **independence claims** the DAG
implies: every pair of non-adjacent nodes should be conditionally independent
given their parents. `basis_set()` lists those claims.

```{r basis}
basis_set(sem)
```

In `drmSEM`, d-separation uses the **any-component LRT**. A missing arrow
`X -> Y` asserts that `X` has *no* effect on *any* modelled component of `Y`. We
test it by refitting `Y`'s node augmented with `X` in each component sub-model
and comparing to the baseline by a likelihood-ratio test:

`LRT = 2 * (logLik(augmented) - logLik(baseline))`,
distributed (asymptotically) as chi-squared with degrees of freedom equal to the
number of added coefficients across components. `dsep()` runs this for every
claim in the basis set.

```{r dsep}
dsep(sem)
```

`fisher_c()` combines the per-claim p-values into Fisher's C statistic,
`C = -2 * sum(log(p_i))`, with `2k` degrees of freedom for `k` claims. A small
p-value flags a missing path -- including a missing path into a *non-mean*
component, which a mean-only SEM tool could never detect.

```{r fisherc}
fisher_c(sem)
```

## Decomposing effects by simulation

Coefficient products are *not* a valid way to combine paths across non-Gaussian
links or across distributional components. `drmSEM` therefore computes effects
by **Monte-Carlo propagation** over the fitted DAG: it sets the exogenous
predictor to two contrasting values, pushes the change through every downstream
node, and reads the population-average change in the response-scale mean of the
target.

**Direct (controlled) effect.** Hold all mediators at their observed values;
move only `temp`. Only the arrows that go *directly* into `survival` operate.

```{r direct}
direct_effects(sem, from = "temp", to = "survival")
```

**Total effect.** Let every mediator respond. With
`mediation = "distribution"`, mediators pass *realized draws* from their fitted
families, so effects flowing through a mediator's `sigma`, `zi`, or `nu` are
included. With `mediation = "mean"`, only mediator means propagate.

```{r total}
total_effects(sem, from = "temp", to = "survival", mediation = "distribution")
```

**Indirect effect with a distributional decomposition.** This is the heart of
the package. `indirect_effects()` reports five quantities:

- `total_path` -- the full simulated effect through the chosen mediators;
- `direct` -- the controlled direct effect;
- `indirect` -- `total_path - direct`;
- `mean_mediated` -- the part carried by mediator *means*;
- `distribution_mediated` -- the *extra* effect that appears only when mediators
  pass realized draws, i.e. the part flowing through scale, zero-inflation, or
  shape (`total under distribution` minus `total under mean`).

```{r indirect}
indirect_effects(sem, from = "temp", to = "survival")
```

A non-trivial `distribution_mediated` row is the signature of this framework:
temperature reaches survival partly because it changes the *spread* of size and
(via habitat structure) the *zeros* of abundance, not only their means. A
mean-only SEM would report that channel as zero or silently fold it into the
mean. `drmSEM` keeps it visible and correctly labelled.

## Standardizing

When you need comparable path strengths, `standardize()` rescales effects (for
example `method = "latent"` or a range/SD-based scaling) while preserving the
component labels.

```{r standardize}
standardize(sem)
```

## Recap

- `drmSEM` is a **layer** on the `drmTMB` **engine**; one `drmTMB` fit per
  endogenous node, DAG-only, observed variables only.
- Paths are **component-labelled** (`mu`, `sigma`, `nu`, `zi`, `hu`,
  `sd(group)`, `rho12`); a non-`mu` path is never a mean effect.
- d-separation is the **any-component LRT**, combined by **Fisher's C**.
- Direct / indirect / total effects are **simulation-based**, never coefficient
  products, and the indirect decomposition exposes **distribution-mediated**
  pathways explicitly.

For the environment and CI setup that compiles and installs the `drmTMB`
engine, see `CLOUD.md`; for the operating contract, see `AGENTS.md`.
