---
title: "drmSEM overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{drmSEM overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# drmSEM never fits its own likelihoods: every endogenous node is one drmTMB
# fit, and drmTMB compiles TMB (C++) from source. The code chunks below are
# illustrative only and are never executed: this vignette has no simulate step,
# so it defines no `dat` or `phy`, and `eval = has_engine` keeps the chunks
# from running. For runnable, end-to-end examples see the companion vignettes
# (`drmSEM.Rmd`, `phylogenetic-sem.Rmd`).
has_engine <- FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_engine
)
library(drmSEM)
```

This vignette is the map of `drmSEM`: it ties together the engine/layer split,
component-labelled paths, the simulation-based effect calculus, d-separation, and
phylogenetic distributional SEM, and closes with a roadmap. It is deliberately
prose-first; each section points to a companion vignette where the topic is
worked in full.

- *Distributional piecewise SEM with drmSEM* (`drmSEM.Rmd`) — the end-to-end walkthrough.
- *Effect decomposition* (`effect-decomposition.Rmd`) — the simulation calculus in depth.
- *Phylogenetic SEM* (`phylogenetic-sem.Rmd`) — shared ancestry inside a distributional SEM.
- *Comparison with other tools* (`comparison.Rmd`) — where `drmSEM` sits in the landscape.

## The one idea

Most structural equation modelling treats a path `X -> Y` as a statement about
the **expected value** of `Y`. But a predictor can do more than move a mean. It
can change how *variable* a response is, how *over-dispersed* it is, how often it
is a *structural zero*, how much *among-group heterogeneity* it carries, or how
strongly two responses *co-vary in their residuals*. In ecology and evolution
these are real, distinct hypotheses:

- "Warmer sites make body size *more variable*" — a claim about `sigma`, not `mu`.
- "Habitat raises the *probability of absence*, not the mean count" — a claim
  about `zi`.
- "Denser populations reduce *among-site heterogeneity* in survival" — a claim
  about `sd(site)`.

`drmSEM` makes those targets first-class causal claims, keeps them labelled as
such everywhere, and propagates their downstream consequences honestly.

## The engine / layer split

`drmSEM` is a **layer**, not a fitting engine.

- [`drmTMB`](https://github.com/itchyshin/drmTMB) is the **engine**: it fits one
  distributional regression — mean, scale, shape, zero-inflation, hurdle,
  random-effect scale, residual correlation — per response.
- `drmSEM` is the **graph / SEM / d-separation / path / effect-decomposition
  layer**. It *never* fits its own likelihoods.

The consequences are concrete: each endogenous node is **exactly one `drmTMB`
fit**, the system is **piecewise** (a graph of separately fitted local models, in
the `piecewiseSEM` tradition), and the directed graph must be a **DAG** — cycles
are an error. Every assumption about the engine's return shapes lives in a single
adapter, so `drmSEM` re-uses, never re-implements, `drmTMB`'s likelihoods, links,
and family samplers.

There are two ways in, producing the identical object:

```{r build}
# Declarative: one drm_node() per endogenous response, a shared data frame.
sem <- drm_sem(
  size      = drm_node(drmTMB::bf(size ~ temp + habitat, sigma ~ temp),
                       family = stats::gaussian()),
  abundance = drm_node(drmTMB::bf(abundance ~ size + temp, zi ~ habitat),
                       family = drmTMB::nbinom2()),
  survival  = drm_node(drmTMB::bf(cbind(alive, dead) ~ abundance + size),
                       family = drmTMB::beta_binomial()),
  data = dat
)

# Assembler: if you have already fitted the nodes yourself, drm_psem() wires
# existing drmTMB objects into the same drm_sem object — no refitting.
```

This canonical chain, `size -> abundance -> survival`, recurs across the
vignettes. Temperature is exogenous and reaches survival only indirectly, through
the *mean*, the *scale*, and the *zero-inflation* of the intermediate nodes —
exactly the situation that breaks coefficient-product mediation.

## Component-labelled paths

A `drmSEM` edge is the tuple `(from, to, component, link, term)`. The `component`
is the distributional parameter of the target node that the predictor acts on:

| Component  | Symbol      | What a path to it means                          |
| ---------- | ----------- | ------------------------------------------------ |
| mean       | `mu`        | shifts the expected response                     |
| scale      | `sigma`     | changes residual scale / dispersion, not the mean|
| shape      | `nu`        | changes a shape parameter (e.g. tail/skew)       |
| zero-infl. | `zi`        | changes the probability of a structural zero     |
| hurdle     | `hu`        | changes the probability of crossing the hurdle   |
| RE scale   | `sd(group)` | changes among-group heterogeneity                |
| residual r | `rho12`     | changes the residual correlation between the two responses of a bivariate node |

`paths()` returns one row per fitted fixed-effect coefficient across all nodes,
each tagged with its component and link. In the canonical model the `temp` path
into `size` appears twice — once on `mu`, once on `sigma` — and they are not the
same claim. The cardinal rule, enforced in `paths()`, `plot()`, the d-separation
test, and the effect decomposition, is that **a path to a non-`mu` component is
never reported as a mean effect.**

The `rho12` component is the **residual correlation** between the two responses
of a bivariate node (`eps_y1 <-> eps_y2`): a path `rho12 ~ x` says `x` changes
that residual coupling, *not* the mean and *not* a directed `y1 -> y2` arrow.
Targeting `rho12` today requires a bivariate `drmTMB` fit (e.g.
`family = c(gaussian(), gaussian())`, components `mu1`/`mu2`/`sigma1`/`sigma2`/`rho12`)
supplied via `drm_psem()`, after which the `rho12` path is extracted and labelled
like any other component path. The covariance-edge grammar now ships:
`covary()` declares a residual or higher-level covariance edge,
`covariances()` reports those edges separately from `paths()`, and d-separation
knows two covarying responses should not get a `y1 _||_ y2` claim. The
bivariate-node declaration grammar now also ships: `drm_pair()` declares a joint
two-response node, `rho12()` / `corpairs()` report the declared edges, and
`plot()` draws double-headed (residual) / dashed (higher-level) arcs. Only the
joint bivariate *fit* that reads a fitted `rho12` / `corpair` correlation back (a
non-`NA` estimate) remains a roadmap item (see
`docs/design/07-bivariate-covariance-edges.md`).

```{r paths}
paths(sem)   # component-labelled path table
plot(sem)    # the DAG, with sigma / zi edges drawn distinctly from mu edges
```

## The effect calculus: counterfactual, not coefficient products

The classical "indirect = product of path coefficients" identity is exact only on
a single linear, identity-link scale. `drmSEM` paths cross links (a `log`-link
mean feeding a `logit`-link mean downstream) and flow through non-mean
components, where a product of coefficients has no defined response-scale
meaning. So `drmSEM` *simulates* the intervention instead: it sets the predictor
to two contrasting values, pushes the change through every downstream node in
topological order (random effects held at zero), and reads off the
population-average change in the response-scale target. This is counterfactual,
`do()`-style propagation in the tradition of Pearl and of Imai, Keele and
Yamamoto. By default the `direct` quantity is a **controlled direct effect**
(mediators held at their observed values), and the total/indirect split is
simulation-based; these coincide with natural direct and indirect effects only
under linearity with no exposure-mediator interaction. When they diverge,
`indirect_effects(effect = "natural")` returns the cross-world **natural**
decomposition (`natural_direct`, `natural_indirect`, `mediated_interaction`)
instead.

**Mean vs distribution mediation.** What an active mediator passes downstream is
the crux. Under `total_effects(method = "gcomp")` it passes its expected `mu`;
under `method = "simulate"` it passes a **realized draw** from its fitted family,
using all of its response-scale parameters (`sigma`, `zi`, `nu`, trials, ...).
Only the latter carries signal that lives in a mediator's scale, zeros, or shape.

**The decomposition.** `indirect_effects()` reports five quantities:

- `total_path` — the full simulated effect with mediators active;
- `direct` — the controlled direct effect (mediators held inactive);
- `indirect` — `total_path - direct`;
- `mean_mediated` — the part carried by mediator *means*;
- `distribution_mediated` — the *extra* effect that appears only when mediators
  pass realized draws, i.e. signal flowing through scale, zero-inflation, or
  shape (distribution total minus mean total).

So `indirect` is approximately `mean_mediated + distribution_mediated`, and the
`distribution_mediated` row isolates exactly the contribution that no
coefficient-product method can express. The default `direct` is a controlled
direct effect, which coincides with the natural direct and indirect effects only
under linearity with no exposure-mediator interaction. For the general case,
`indirect_effects(effect = "natural")` returns the cross-world natural
decomposition — `natural_direct`, `natural_indirect`, and
`mediated_interaction` (Pearl; Imai, Keele and Yamamoto) — validated against the
linear-Gaussian recovery case.

```{r effects}
direct_effects(sem,   from = "temp", to = "survival")
total_effects(sem,    from = "temp", to = "survival", method = "simulate")
indirect_effects(sem, from = "temp", to = "survival")  # the five-row decomposition
indirect_effects(sem, from = "temp", to = "survival", effect = "natural")  # NDE / NIE / mediated interaction
```

Note: `direct_effects()` reports the mean (response) scale only. Non-mean
channels (`sigma`, `zi`, ...) do not appear here — they show up in `paths()` and
in the `distribution_mediated` row of `indirect_effects()`.

**Outcome functionals.** Because effects come from simulated populations, the
target need not be a mean. `total_effects(target = ...)` reports the
distribution-mediated effect on a response-scale functional — `Pr(Y > t)`
(`"p_gt"`), the zero probability `Pr(Y = 0)` (`"p_zero"`), or the variance of
`Y` (`"var"`) — which is often the scientifically relevant quantity and is the
headline estimand of distributional SEM. These are recovery-tested (for example,
the `p_zero` effect recovers the Poisson zero-probability change). Coefficient
uncertainty is propagated by drawing each node's coefficients from
`MVN(coef, vcov)` and repeating the whole propagation. See
`effect-decomposition.Rmd` for the full treatment.

```{r functionals}
total_effects(sem, from = "temp", to = "survival", target = "p_gt", threshold = 0)
total_effects(sem, from = "temp", to = "survival", target = "p_zero")
total_effects(sem, from = "temp", to = "survival", target = "var")
```

## d-separation: an any-component test

A piecewise SEM is tested by the **independence claims** its DAG implies: every
pair of non-adjacent nodes should be conditionally independent given their
parents. `drmSEM` sharpens the usual mean-only version. Because each node models a
whole conditional distribution, a missing arrow `X -> Y` asserts that `X` predicts
**none** of `Y`'s modelled components — `mu`, `sigma`, `zi`, and the rest.
Adjacency is correspondingly any-component: an arrow into `zi(Y)` alone makes `X`
adjacent to `Y`.

`basis_set()` enumerates the claims. `dsep()` tests each by refitting `Y`'s node
with `X` added to *every* component sub-formula and comparing to the base node by
a likelihood-ratio test, with degrees of freedom equal to the number of added
coefficients. `fisher_c()` combines the claim p-values into Fisher's C,
`C = -2 * sum(log p)`, on `2k` degrees of freedom for `k` claims; a small p-value
flags a missing path — including a missing path into a *non-mean* component, which
a mean-only tool could never detect.

```{r dsep}
basis_set(sem)   # the implied independence claims
dsep(sem)        # any-component LRT per claim
fisher_c(sem)    # combined goodness-of-fit
```

The any-component definition is a deliberate, novel research choice: it treats "X
is irrelevant to Y" as a statement about Y's whole conditional distribution. Its
OQ-6 calibration grid (mean-only, distributional, and cross-link DGPs; `n` from
100 to 1000; 200 replicates per cell) met the pre-specified Type-I,
augmented-component-count, Fisher-uniformity, and power criteria. Broader
Fisher's C and d-separation settings remain claim-scoped until separately tested.
Claims whose augmented refit cannot be formed are flagged and dropped from
Fisher's C rather than silently scored.

## Phylogenetic distributional SEM

Shared ancestry enters a `drmSEM` model as a structured random effect on a node.
`drmTMB` exports the markers `phylo()`, `animal()`, `relmat()`, and `spatial()`,
and `drmSEM` recognises them as markers: they are stripped from the causal edge
set, so a node can carry a phylogenetic random effect *today* while `paths()`,
`dsep()`, `fisher_c()`, and the effect engine operate on the fixed-effect DAG.

```{r phylo}
sem_phylo <- drm_sem(
  y1 = drm_node(drmTMB::bf(y1 ~ x + phylo(1 | species, tree = phy)),
                family = stats::gaussian()),
  y2 = drm_node(drmTMB::bf(y2 ~ y1 + x + phylo(1 | species, tree = phy)),
                family = stats::gaussian()),
  data = dat
)
paths(sem_phylo)
indirect_effects(sem_phylo, from = "x", to = "y2")
```

This is the genuinely new contribution. Existing phylogenetic SEM tools
(`phylopath`, `phylosem`) focus on causal relationships among trait *means* under
phylogenetic covariance; `drmSEM` extends this to **distributional** SEM, where
shared ancestry can be incorporated while direct and indirect effects are
estimated for changes in means, variances, zero probabilities, and other
response-distribution components. Practical notes: `phy` is an `ape` "phylo"
object and must be ultrametric (rescale a raw tree with
`ape::compute.brlen(phy, "Grafen")`), and every `species` level must be a tip
label. Phylopath-style model-set comparison (`drm_dag()` / `drm_model_set()` /
`compare()` / `best()` / `average()`, Phase 2) and evolutionary covariance
models (`drm_phylo_cov()` for BM / lambda / OU / kappa on a *fixed* grid,
Phase 3) ship today; phylogenetic paths *into* distributional components and
*jointly estimated* evolutionary covariance remain roadmap items. See
`phylogenetic-sem.Rmd`.

## Roadmap

| Capability | Status |
| --- | --- |
| Component-labelled paths (`mu`, `sigma`, `nu`, `zi`, `hu`, `sd(group)`, `rho12`) | now (a `x -> rho12` path requires a bivariate `drmTMB` fit via `drm_psem()`) |
| Bivariate covariance-edge grammar: `covary()` declaration + `covariances()` accessor + covariance-aware d-separation (drops the `y1 _||_ y2` claim) | now |
| Bivariate-node *declaration*: `drm_pair()` grammar, `rho12()` / `corpairs()` accessors (declared edges), double-headed / dashed-arc plotting | now (0.2.x) |
| Bivariate joint *fit*: `rho12()` / `corpairs()` read-back of a *fitted* (non-`NA`) correlation from a live bivariate `drmTMB` fit | roadmap (0.4 engine) |
| Any-component d-separation + Fisher's C | now (OQ-6 calibration grid validated; broader settings claim-scoped) |
| Simulation-based direct / mean-mediated / distribution-mediated / total effects | now |
| Natural direct / indirect / mediated-interaction effects (`indirect_effects(effect = "natural")`) | now (cross-world split, validated on the linear-Gaussian case) |
| Outcome functionals beyond the mean (`Pr(Y>t)`, `Pr(Y=0)`, var via `direct_effects()` / `total_effects(target = ...)`) | now (first set; more functionals + bootstrap CIs on roadmap) |
| Unified effect-API surface (`method`, `uncertainty`, `nsim`, `population` shared across the effect functions) | now (`mediation` / `draw` / `n_sim` kept as deprecated aliases; bootstrap + marginal are roadmap) |
| Path-specific attribution: per-mediator (`inclusion`/`exclusion`), per-component (`mean`/`sigma`/`zi`), and natural variant with recanting-witness flag (`path_effects()`) | now |
| Composite (formative / PCA) latent constructs with reliability + loadings (`drm_composite()` / `loadings()`) | now |
| Component-labelled DAG plot + effect forest plot | now |
| Phylogenetic correction inside piecewise SEM (Phase 1, intercept-only) | now |
| Phylopath-style model-set comparison (best / averaged model) | now (Phase 2; `drm_dag` / `drm_model_set` / `compare` / `best` / `average`) |
| Evolutionary covariance models (BM / OU / Pagel's lambda / kappa) | now (Phase 3; `drm_phylo_cov`, fixed grid — not jointly estimated) |
| Jointly estimated evolutionary covariance (λ / OU / κ) | roadmap |
| Phylogenetic paths *into* distributional components | roadmap (Phase 4) |
| Reflective latent measurement models (latent estimated jointly with the structural model) | out of scope for 0.x (needs a joint likelihood; composite constructs ship — see above) |
| Full joint (single-likelihood) multivariate SEM | out of scope for 0.x |
| Declared feedback motifs: `drm_cycle()` / `feedback=` grammar, relaxed toposort, d-sep suppression, internal equilibrium propagator | now (0.5.0; equilibrium effects via the public effect API are the 0.5.x increment) |
| Undeclared cyclic / feedback graphs | an error (declare the motif with `drm_cycle()` to allow it) |
| New likelihoods | out of scope (drmSEM never fits its own) |

## Recap

- `drmSEM` is a **layer** on the `drmTMB` **engine**: one fit per endogenous
  node, DAG-only, observed variables only.
- Paths are **component-labelled**; a non-`mu` path is never a mean effect.
- d-separation is the **any-component LRT**, combined by **Fisher's C**.
- Effects are **counterfactual and simulation-based**, never coefficient
  products, and the indirect decomposition exposes the **distribution-mediated**
  channel explicitly.
- Phylogeny enters as a structured random effect, giving a first slice of
  **phylogenetic distributional SEM**.

For the full walkthrough see `drmSEM.Rmd`; for effects, `effect-decomposition.Rmd`;
for phylogeny, `phylogenetic-sem.Rmd`; for positioning, `comparison.Rmd`.
