---
title: "Effect decomposition"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Effect decomposition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# drmSEM never fits its own likelihoods: every endogenous node is one drmTMB
# fit, and drmTMB compiles TMB (C++) from source. While the drmTMB-integration
# layer is still pending end-to-end validation (see
# docs/memory/VALIDATION_LEDGER.md), the model-fitting and effect chunks below
# are shown for illustration but not evaluated. Flip `has_engine` to
# requireNamespace("drmTMB", quietly = TRUE) once integration is validated.
has_engine <- FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_engine
)
library(drmSEM)
```

## The picture no other SEM package draws

Every structural equation modelling tool will draw you a path diagram. None of
them will draw you *this*: a forest plot that splits the total effect of a
predictor into the part that travels through mediator **means** and the part
that travels through mediator **spread, zeros, or shape**. That second channel
-- the **distribution-mediated** effect -- is the reason `drmSEM` exists, and it
is the figure below.

```{r headline, eval = requireNamespace("ggplot2", quietly = TRUE), echo = FALSE, fig.width = 7, fig.height = 3.2, fig.alt = "Forest plot decomposing the effect of size on fitness into total, direct, indirect, mean-mediated and distribution-mediated quantities, with the distribution-mediated channel a clear non-zero slice."}
# Built BY HAND so the marquee figure renders without the drmTMB engine.
# These are illustrative numbers for the size -> abundance -> fitness chain
# (see below); the distribution-mediated row is a clear non-zero slice of the
# indirect effect.
eff <- data.frame(
  from = "size", to = "fitness",
  quantity = c("total_path", "direct", "indirect",
               "mean_mediated", "distribution_mediated"),
  estimate  = c(0.62, 0.20, 0.42, 0.27, 0.15),
  conf.low  = c(0.41, 0.06, 0.24, 0.13, 0.04),
  conf.high = c(0.83, 0.34, 0.60, 0.41, 0.26),
  stringsAsFactors = FALSE
)
class(eff) <- c("drm_effect", "data.frame")
plot(eff)
```

In this **illustrative** figure (the numbers above are hand-set, not estimated
from data) the orange `distribution_mediated` interval sits clear of zero. That
is the picture to look for: a part of how `size` reaches `fitness` flowing
through the *variability* and *zero-structure* of an intermediate count rather
than through any mean. A mean-only SEM reports that channel as zero, or silently
folds it into the mean. `drmSEM` isolates it and labels it.

## The three effect types

`drmSEM` distinguishes three ways a predictor `X` can reach a target node `Y`.
The equations below (from `docs/design/02-effect-calculus.md`) name the exact
component path each one travels.

- **Direct.** `X -> mu(Y)`. The controlled direct effect: hold every mediator at
  its observed value and move only `X`, so just the arrows running *directly*
  into the mean of `Y` operate.

- **Mean-mediated.** `X -> mu(M1) -> mu(M2) -> mu(Y)`. The classical indirect
  path: `X` shifts the *mean* of a mediator, which shifts the *mean* of the next,
  on down to the mean of `Y`.

- **Distribution-mediated.** `X -> sigma(M2) -> distribution(M2) -> mu(Y)`. The
  part of the indirect effect that flows through a mediator's *scale* (`sigma`),
  *zero-inflation* (`zi`), or *shape* (`nu`) rather than its mean. `X` changes
  how *spread out* (or how *zero-inflated*, or how *skewed*) the mediator is;
  that altered distribution, pushed through a **curved** downstream outcome,
  moves the mean of `Y`. Concretely it is a **Jensen gap** — the difference
  between propagating the mediator's full distribution and propagating only its
  mean. It is therefore non-zero only when **both** a higher moment of the
  mediator responds to `X` **and** the outcome is nonlinear in that mediator: it
  is exactly zero under mean mediation, and zero when the outcome is *linear* in
  `M` even if `sigma(M)` depends on `X`.

That third type is what makes `drmSEM` distinctive: it has **no
coefficient-product analogue**, because the signal lives in a distributional
parameter and only becomes a mean effect after passing through a nonlinear link
downstream — so a mean-only SEM reports it as zero. The *estimand* itself is not
new: it is the interventional / distributional-mediation effect of the causal
mediation literature (Pearl; Imai, Keele & Yamamoto; VanderWeele; Vansteelandt &
Daniel). What is new is making a mediator's `sigma`/`zi`/shape a first-class
causal target in a piecewise distributional GLMM SEM and surfacing this term as
a labelled, uncertainty-quantified row. It is identified under the usual
mediation assumptions **plus** a correctly specified mediator *distribution*
(its higher moments, not just its mean).

## Why coefficient products are invalid here

The textbook mediation identity "indirect = product of path coefficients" is
exact only on a **single linear, identity-link scale**. A `drmSEM` path routinely
breaks both conditions:

- **It crosses links.** A `log`-link Poisson mean can feed a `logit`-link
  binomial mean downstream. Multiplying a coefficient defined on the log scale by
  one defined on the logit scale produces a number with no response-scale
  meaning.

- **It crosses distributional components.** A `temp -> sigma(size)` coefficient
  describes residual *spread*, not the *mean*. There is no valid arithmetic that
  multiplies a spread coefficient by a downstream mean coefficient to get a mean
  effect.

So `drmSEM` does not multiply coefficients. It **simulates the intervention**:
set the predictor to two contrasting values (a one-SD contrast for numerics, the
first two levels for factors), push each contrast through the whole fitted DAG
with random effects held at zero, and read off the population-average change in
the response-scale mean of the target. The distribution-mediated effect is
*only expressible at all* under this simulation view -- it has no
coefficient-product counterpart to compare against.

## The five-number decomposition

`indirect_effects()` returns five rows, one per `quantity`. Each is a contrast of
population-average response means between the high and low intervention.

| `quantity` | meaning |
| --- | --- |
| `total_path` | the full simulated effect with mediators active and passing realized draws |
| `direct` | the controlled direct effect (mediators held inactive) |
| `indirect` | `total_path - direct` |
| `mean_mediated` | (mean-mediated total) `-` direct: the part carried by mediator *means* |
| `distribution_mediated` | (distribution total) `-` (mean total): the *extra* part flowing through mediator `sigma` / `zi` / `nu` |

The identity to keep in mind is

```
indirect  ~=  mean_mediated  +  distribution_mediated
```

Reading the headline forest plot top to bottom: `total_path` is everything;
`direct` is the slice that bypasses the mediators; `indirect` is what is left
once the direct slice is removed; and that indirect slice splits into the blue
`mean_mediated` part and the orange `distribution_mediated` part. When the orange
interval excludes zero, you have evidence that the predictor acts on a
mediator's distribution shape, not just its location.

## The canonical example: size -> abundance -> fitness

The data-generating process behind the headline figure (see the `drmSEM`
introductory vignette and `tests/testthat/helper-dgp.R` for the full version) is
an ecological chain where `size` reaches `fitness` through an intermediate count,
`abundance`. Crucially, `size` is wired into a **non-mean** component of
`abundance`: it inflates the *spread* / *zero-structure* of the count, not only
its mean. That is what populates the distribution-mediated row.

The model is built and queried exactly as below. These chunks call the engine on
a real fit, so they are shown but not run here.

```{r build, eval = has_engine}
# size      -> abundance (count): size acts on BOTH the NB mean and its
#              dispersion/zero-inflation, so it has a distributional channel.
# abundance -> fitness  (proportion alive out of trials).
sem <- drm_sem(
  abundance = drm_node(
    drmTMB::bf(abundance ~ size + temp, zi ~ size),   # size -> zi(abundance)
    family = drmTMB::nbinom2()
  ),
  fitness   = drm_node(
    drmTMB::bf(cbind(alive, dead) ~ abundance + size),
    family = drmTMB::beta_binomial()
  ),
  data = dat
)
```

`paths()` shows the component-labelled coefficients; note the `size` path into
`abundance` appears once on `mu` and once on `zi` -- two different claims.

```{r paths, eval = has_engine}
paths(sem)
```

The controlled direct effect moves only `size` and lets nothing mediate:

```{r direct, eval = has_engine}
direct_effects(sem, from = "size", to = "fitness")
```

The total effect under distribution mediation lets `abundance` pass *realized
draws* from its negative-binomial family, so the spread and zeros it carries are
included:

```{r total, eval = has_engine}
total_effects(sem, from = "size", to = "fitness", method = "simulate")
```

And the full five-number decomposition -- the object plotted at the top of this
vignette -- comes from a single call:

```{r indirect, eval = has_engine}
eff <- indirect_effects(sem, from = "size", to = "fitness")
eff          # the five quantities as a table
plot(eff)    # the forest plot
```

`plot()` on the returned `drm_effect` draws the same forest plot rendered above:
black for direct / total, blue for the mean-mediated channel, orange for the
distribution-mediated channel, with a dashed reference line at zero.

## Emphasis: the distribution-mediated channel

It is worth being explicit about what makes the orange row non-zero. In this
example `size` enters `zi(abundance)` -- the *probability that a count is a
structural zero*. Larger organisms are found in patches that are present-or-
absent more sharply, so `size` reshapes the zero-structure of `abundance`
without (necessarily) moving its mean. When those realized, zero-heavy counts
are pushed through the `logit`-link survival model, they move mean fitness. That
movement is the distribution-mediated effect.

The same logic applies to any non-mean component:

- a path into `sigma(M)` -- `X` makes a mediator more or less *variable*;
- a path into `zi(M)` or `hu(M)` -- `X` changes the *zeros* or the *hurdle*;
- a path into `nu(M)` -- `X` changes a *shape* parameter (tails / skew).

In each case the effect on `Y` is real **provided the downstream outcome is
curved in that mediator** (a linear outcome has no Jensen gap, so the row is
zero even when the mediator's distribution responds to `X`); it is zero under
mean mediation, and is reported on its own row, correctly labelled, by
`indirect_effects()`. No other SEM package plots it as a separate row -- which is
exactly why this vignette opens with the figure that does.

## Recap

- Three effect types: **direct** (`X -> mu(Y)`), **mean-mediated**
  (`X -> mu(M) -> mu(Y)`), and **distribution-mediated**
  (`X -> sigma(M) -> distribution(M) -> mu(Y)`).
- Effects are **simulation-based**, never coefficient products, because paths
  cross links and distributional components.
- `indirect_effects()` returns a five-number decomposition; the
  `distribution_mediated` row isolates the channel no coefficient-product method
  can express.
- `plot()` on the result is the marquee figure: a forest plot that makes the
  distributional channel visible.

For the full three-node worked system, d-separation, and Fisher's C, see the
`drmSEM` introductory vignette.
