---
title: "Equations via symbolizer"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Equations via symbolizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# drmSEM never fits its own likelihoods. The integration with symbolizer
# operates on already-fitted nodes (each one a drmTMB fit), so this vignette
# is illustrative and not evaluated when drmTMB or symbolizer is missing.
has_engine <- requireNamespace("drmTMB", quietly = TRUE) &&
              requireNamespace("symbolizer", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_engine)
library(drmSEM)
if (has_engine) {
  set.seed(13)
  n <- 300
  temp <- stats::rnorm(n)
  habitat <- factor(sample(c("A", "B"), n, replace = TRUE))
  habitat_b <- habitat == "B"
  size <- stats::rnorm(
    n,
    mean = 0.3 + 0.6 * temp + 0.4 * habitat_b,
    sd = exp(-0.3 + 0.3 * temp)
  )
  mu_abundance <- exp(0.5 + 0.25 * size + 0.15 * temp)
  zi_abundance <- stats::plogis(-1 + 0.8 * habitat_b)
  abundance <- stats::rnbinom(n, mu = mu_abundance, size = 2)
  abundance[stats::runif(n) < zi_abundance] <- 0L
  dat <- data.frame(temp, habitat, size, abundance)
}
```

`drmSEM` produces a **graph of distributional models**: each endogenous node is
one `drmTMB` fit, and paths into `mu` / `sigma` / `nu` / `zi` / `hu` / `rho12`
are first-class. Reporting that kind of model in a paper requires writing the
per-node equations, families, links, and assumptions — error-prone work that
also goes stale when the code changes.

[`symbolizer`](https://itchyshin.github.io/symbolizer/) renders the equations
**from the fitted object** itself. Its `drmTMB` extractor covers exactly the
families and components `drmSEM` uses, so wiring it up is a thin bridge: walk
the SEM in topological order, symbolize each node's `drmTMB` fit, and collate.

## A two-node example

```{r build}
sem <- drm_sem(
  size      = drm_node(drmTMB::bf(size ~ temp + habitat, sigma ~ temp),
                       family = stats::gaussian()),
  abundance = drm_node(drmTMB::bf(abundance ~ size + temp, zi ~ habitat),
                       family = drmTMB::nbinom2()),
  data = dat
)
```

## Render

```{r symbolize}
sym <- symbolizer::symbolize(sem) # walks sem$records, symbolizes each node
sym                          # prints a summary header

symbolizer::as_latex(sym)    # publication-quality equations, node-by-node
symbolizer::equations(sym)   # tidy tibble: one row per component, keyed by node
symbolizer::assumption_table(sym) # families, links, deferred components
```

The collator labels each block with `## Node: <name>` so the reader can see
exactly which equations belong to which response. `sigma` and `zi` sub-models
appear next to their parent `mu`, with the link function spelled out — which
makes the component-labelled grammar of `drmSEM` concrete instead of abstract.

## Scope

`symbolize.drm_sem()` covers the **model specification** layer (per-node
distributional equations, families, links, assumptions). The **SEM layer** —
the DAG, [`dsep()`], [`fisher_c()`], and the simulation-based effect calculus —
stays with `drmSEM` itself, because the **distribution-mediated effect** row
of [`indirect_effects()`] has no closed-form symbolic representation by
construction. Symbolizer correctly stops at the per-node equations and lets
`drmSEM` do the propagation.

## See also

- `vignette("effect-decomposition")` — the simulation-based effect calculus
  that lives on top of these node equations.
- `vignette("drmSEM-overview")` — the broader workflow.
- The [`symbolizer` site](https://itchyshin.github.io/symbolizer/) — supported
  families and components, equation rendering options, and the
  `symbolized_model` object.

## References
