---
title: "Feedback cycles: reciprocal causation and equilibrium effects (drm_cycle)"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Feedback cycles: reciprocal causation and equilibrium effects (drm_cycle)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Model-fitting chunks use the drmTMB engine and are eval-gated; the pure-R
# declaration grammar (drm_cycle, cycles) runs.
has_engine <- requireNamespace("drmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
library(drmSEM)
set.seed(20260828)
```

In standard piecewise structural equation models [@Shipley2000; @Shipley2009;
@Lefcheck2016], causal dependencies flow in one direction along a **directed
acyclic graph (DAG)**. However, many ecological, behavioural, and physiological
systems feature **reciprocal feedback loops**—where two variables simultaneously
cause each other:

- **Physiology & Behaviour**: An animal's glucocorticoid stress levels suppress its movement, while high activity elevates physiological stress (`activity ⇄ stress`).
- **Ecosystems**: Predator and prey densities mutually regulate one another (`predator ⇄ prey`).
- **Plant–Soil Interactions**: Plant biomass enriches microbial activity, which in turn enhances nutrient uptake and plant growth (`plant ⇄ soil`).

Such feedback breaks the topological ordering assumed by DAG algorithms. This
vignette explains how `drmSEM` enables researchers to:

1. **Declare feedback motifs** explicitly using `drm_cycle()`.
2. Compute **equilibrium total effects** via **Banach fixed-point iteration** across location ($\mu$), scale ($\sigma$), shape ($\nu$), and zero-inflation ($\text{zi}$) components.
3. Diagnose system stability using the **spectral radius $\rho(B) < 1$** and empirical **Lipschitz contraction constants**.
4. Understand the honest boundary between simulation-based equilibrium calculation and simultaneity bias in node-wise estimation.

---

## The reciprocal challenge: DAGs versus cycles

A reciprocal pair introduces two distinct statistical challenges:

1. **Simultaneity bias during fitting**:
   In $y_1 \rightleftarrows y_2$, $y_1$ is a predictor in the $y_2$ equation, but $y_1$ is also a descendant of $y_2$. Therefore, $y_1$ correlates with $y_2$'s error term. Ordinary node-wise maximum likelihood is **inconsistent** [@Bollen1989; @Bollen1996].
2. **Equilibrium propagation**:
   In a DAG, the total effect of an upstream driver $x$ is computed by propagating changes down a finite sequence of directed paths (topological sweep). In a cyclic system, there is no topological order: an intervention reverberates around the loop indefinitely until the system relaxes to an **equilibrium steady state**.

---

## 1. Declaring a feedback motif: `drm_cycle()`

By default, `drmSEM` enforces strict acyclicity: any undetected cycle results in
a **hard error**. To permit feedback, you must explicitly declare the motif using
`drm_cycle()`:

```{r declare-cycle}
# Declare a reciprocal pair between activity and stress
motif <- drm_cycle("activity", "stress")
motif
```

When handed to `drm_sem(..., feedback = motif)` or `drm_psem(..., feedback = motif)`:
- The topological sort condenses the motif into a super-node (relaxed topological sort).
- Independence claims between motif members ($y_1 \perp\mkern-10mu\perp y_2$) are suppressed from the d-separation basis set.
- All undeclared cycles remain strict errors.

```{r sem-fit, eval = has_engine}
# Simulate data with reciprocal feedback
n <- 300
x <- rnorm(n)
# Reduced-form equilibrium simulation:
# activity = 0.4*stress + 0.5*x + e1
# stress   = -0.3*activity + 0.4*x + e2
# B = matrix(c(0, -0.3, 0.4, 0), 2, 2)
# (I - B)^(-1) is well-conditioned because spectral radius rho(B) = sqrt(0.12) = 0.346 < 1
I_minus_B_inv <- solve(diag(2) - matrix(c(0, -0.3, 0.4, 0), 2, 2))
Gamma <- matrix(c(0.5, 0.4), 2, 1)
total_mult <- I_minus_B_inv %*% Gamma

e1 <- rnorm(n, 0, 0.3)
e2 <- rnorm(n, 0, 0.3)
reduced_e <- t(I_minus_B_inv %*% rbind(e1, e2))

dat <- data.frame(
  x = x,
  activity = total_mult[1, 1] * x + reduced_e[, 1],
  stress   = total_mult[2, 1] * x + reduced_e[, 2]
)

# Fit the SEM with declared feedback:
sem_cyclic <- drm_sem(
  activity = drm_node(drmTMB::bf(activity ~ stress + x), family = stats::gaussian()),
  stress   = drm_node(drmTMB::bf(stress ~ activity + x), family = stats::gaussian()),
  data = dat,
  feedback = drm_cycle("activity", "stress")
)

# Inspect declared cycles
cycles(sem_cyclic)
```

---

## 2. Multi-component Banach fixed-point iteration

In linear Gaussian models, the steady-state equilibrium under an intervention
$\Delta x$ has a closed-form matrix solution [@Bollen1989]:

$$y = (I - B)^{-1} \Gamma x$$

where $B$ is the direct within-loop coefficient matrix (with zeros on the diagonal)
and $\Gamma$ contains direct effects of exogenous variables.

Because $(I - B)^{-1} = I + B + B^2 + B^3 + \dots$, the matrix inverse is the
infinite sum over all walks around the loop.

### Distributional and non-linear systems

In distributional regression, responses may follow non-Gaussian families
(e.g., Gamma, Negative Binomial, Beta, Student-$t$) or have location-scale
links ($\log \sigma$, $\text{logit } \text{zi}$, $\log \nu$). For non-linear
systems, no closed-form matrix inverse exists.

`drmSEM` resolves non-linear feedback using **vectorized Banach fixed-point iteration**
[@Banach1922]. Let $\mathbf{s}^{(t)}$ denote the joint state vector containing
the predicted values across all active distributional components:

$$\mathbf{s}^{(t)} = \begin{pmatrix} \mu_{\text{activity}}^{(t)} \\ \sigma_{\text{activity}}^{(t)} \\ \mu_{\text{stress}}^{(t)} \\ \sigma_{\text{stress}}^{(t)} \end{pmatrix}$$

At each iteration $t = 1, 2, \dots, \text{max\_iter}$:
1. Each endogenous node re-predicts all its distributional components from the current working values of its parents.
2. The state vector updates with adaptive relaxation:
   $$\mathbf{s}^{(t+1)} = (1 - \alpha) \mathbf{s}^{(t)} + \alpha \, \mathcal{T}(\mathbf{s}^{(t)})$$
3. The iteration terminates when the Euclidean norm of change is below tolerance:
   $$\|\mathbf{s}^{(t+1)} - \mathbf{s}^{(t)}\|_\infty < \text{tol} \quad (\text{default } 10^{-8})$$

---

## 3. Stability diagnostics and Lipschitz contraction

A feedback system reaches a stable equilibrium if and only if the iteration is a
**strict contraction** [@Banach1922; @ForreMooij2017]. `drmSEM` monitors stability
through two complementary diagnostics:

### 1. Spectral radius $\rho(B)$
For the linear Gaussian sub-system, `drmSEM` computes the eigenvalues of $B$:

$$\rho(B) = \max_i |\lambda_i(B)|$$

- **$\rho(B) < 1$**: The feedback is damped. The series converges to a unique, stable equilibrium.
- **$\rho(B) \ge 1$**: The feedback is explosive / unstable. Trips around the loop amplify without bound, and **no equilibrium effect is defined**.

### 2. Empirical Lipschitz contraction constant
For non-linear and distributional loops, `drmSEM` calculates the empirical
ratio of successive state deltas:

$$L^{(t)} = \frac{\|\mathbf{s}^{(t+1)} - \mathbf{s}^{(t)}\|}{\|\mathbf{s}^{(t)} - \mathbf{s}^{(t-1)}\|}$$

If $L < 1$, the map is contractive on the support of the data.

---

## 4. Equilibrium total effects

To estimate the downstream consequence of an intervention in a feedback system,
call `total_effects()`:

```{r total-effects, eval = has_engine}
eff_total <- total_effects(sem_cyclic, from = "x", to = "stress")
eff_total
```

### Key properties of equilibrium effects

1. **`mediation = "equilibrium"`**:
   The output explicitly flags that the estimate was computed by fixed-point equilibrium simulation rather than a finite acyclic path product.
2. **Honest divergence handling**:
   If a system is unstable ($\rho(B) \ge 1$ or divergence), `drmSEM` returns `NA` with an informative warning rather than a fabricated or infinite number.
3. **Decomposition constraints**:
   Because path-specific mediation decompositions require a topological sort, `indirect_effects()` and `path_effects()` are disabled on cyclic SEMs, directing users to `total_effects()`. `direct_effects()` remains valid (measuring the controlled direct effect holding mediators constant).

---

## 5. d-Separation and basis-set scoping

In acyclic graphs, conditional independence claims are evaluated using
d-separation [@Pearl2009]. In cyclic graphs, the correct independence criterion is
**$\sigma$-separation** [@ForreMooij2017].

Until full $\sigma$-separation is implemented, `basis_set()` automatically
**drops conditional independence claims among members of declared feedback motifs**,
scoping Fisher's C test strictly to the acyclic components of the network.

---

## Summary of best practices

1. **Explicitly declare cycles**: Never let a feedback loop exist undeclared; use `drm_cycle("y1", "y2")`.
2. **Check stability**: Ensure $\rho(B) < 1$ and verify that fixed-point iteration reaches convergence.
3. **Report equilibrium effects**: Use `total_effects()` to capture the fully reverberated steady-state response.
4. **Be honest about simultaneity**: If fitting node-wise regressions, acknowledge that standard ML coefficients in reciprocal loops are subject to simultaneity bias.

---

## References
