---
title: "Latent constructs: composites and reliability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Latent constructs: composites and reliability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Model-fitting chunks use the drmTMB engine and are eval-gated; the pure-R
# construct helpers (drm_composite, print, summary) run.
has_engine <- FALSE
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_engine)
library(drmSEM)
set.seed(1)
```

drmSEM lets a node load on a **latent construct** while staying inside its hard
constraints — observed-variable, piecewise, DAG-only, each node one `drmTMB` fit,
no joint likelihood. The piece that fits this paradigm is the **composite
construct**: a deterministic index of observed indicators, materialized as a
column *before* fitting, so it can be used as an ordinary predictor or response.

## A formative composite

A **formative** construct is *defined* by its indicators (they are causes, not
effects, of the construct). `drm_composite()` builds it; the helper is pure R and
runs without the engine:

```{r formative, eval = TRUE}
dat <- data.frame(
  bill = rnorm(200), tarsus = rnorm(200), mass = rnorm(200), x = rnorm(200)
)
# equal-weighted index of three size indicators:
size <- drm_composite("size", c("bill", "tarsus", "mass"), data = dat)
size
```

Pass it to `drm_sem(..., composites = )`; the column is materialized before
fitting, so a node formula can use `size` like any observed predictor, and
`loadings()` reports the indicator weights (kept separate from `paths()`):

```{r formative-sem}
sem <- drm_sem(
  fitness = drm_node(drmTMB::bf(fitness ~ size + x), family = stats::gaussian()),
  data = dat,
  composites = drm_composite("size", c("bill", "tarsus", "mass"), data = dat)
)
loadings(sem)
```

## A reflective-flavoured composite, and construct quality

When the indicators are thought to be *effects* of one underlying dimension
(reflective), the first principal component approximates that common factor.
`method = "pca"` records the PC1 loadings and the proportion of variance it
captures, and every composite reports its **reliability** (Cronbach's alpha — the
internal consistency of the indicator set):

```{r reflective, eval = TRUE}
# correlated indicators of one latent dimension
z <- rnorm(200)
ind <- data.frame(i1 = z + rnorm(200, sd = .4),
                  i2 = z + rnorm(200, sd = .4),
                  i3 = z + rnorm(200, sd = .4))
boldness <- drm_composite("boldness", c("i1", "i2", "i3"),
                          method = "pca", data = ind, standardize = TRUE)
summary(boldness)
```

A high alpha (indicators move together) supports a reflective reading; a low or
negative alpha is an honest signal that the indicators do *not* form a coherent
scale. `standardize = TRUE` puts the score on a mean-0 / sd-1 scale so it enters a
model comparably to other standardized predictors. (For a *formative* construct
alpha is reported for information only — internal consistency is not the right
criterion when indicators are causes.)

## A construct as a response

Because the construct is just a column, a node can also be fit *on* it — e.g. to
ask what predicts `size`:

```{r response}
sem2 <- drm_sem(
  size    = drm_node(drmTMB::bf(size ~ x), family = stats::gaussian()),
  fitness = drm_node(drmTMB::bf(fitness ~ size), family = stats::gaussian()),
  data = dat,
  composites = drm_composite("size", c("bill", "tarsus", "mass"), data = dat)
)
paths(sem2)
```

## What is *not* a composite

A true **reflective measurement model** — estimating the latent's loadings and
score *jointly* with the structural model, propagating score uncertainty —
requires a single joint likelihood across all indicators, which a piecewise,
one-fit-per-node engine cannot provide. That belongs to the **0.4 joint
multivariate** work (or lavaan interop), and drmSEM does not pretend a composite
is it. A composite is an *observed* index with transparent, recorded weights; use
it when that is what you mean, and reach for a joint factor model when you need
the latent estimated.
