---
title: "Latent constructs and measurement models in drmSEM"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Latent constructs and measurement models in drmSEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Model-fitting chunks use the drmTMB engine and are eval-gated; the pure-R
# construct helpers (drm_latent, drm_indicator, drm_composite, reliability) run.
has_engine <- requireNamespace("drmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
library(drmSEM)
set.seed(20260828)
```

In observational and ecological research, many foundational concepts—such as
**individual quality**, **habitat degradation**, **boldness**, or **organismal
size**—cannot be measured directly with a single error-free variable. Instead,
researchers measure a battery of correlated proxies (indicators) or combine
multiple environmental drivers.

`drmSEM` supports **latent, MIMIC, and composite measurement models** within
its piecewise structural equation modelling framework [@Shipley2000; @Shipley2009;
@Grace2008]. This vignette explains:

1. The distinction between **formative**, **reflective**, and **MIMIC** constructs.
2. The declaration syntax: `drm_latent()`, `drm_indicator()`, and `drm_composite()`.
3. **Identification constraints**: marker indicator ($\lambda_{\text{marker}} = 1$) versus unit-variance ($\text{Var}(\eta) = 1$).
4. **Reliability metrics**: Cronbach's $\alpha$ [@Cronbach1951] versus Raykov's composite reliability $\rho$ [@Raykov1997; @McDonald1999].
5. **Distributional effect propagation**: tracing causal interventions through latent constructs to downstream mean ($\mu$), dispersion ($\sigma$), and zero-inflation ($\text{zi}$) components.
6. The honest piecewise paradigm: how factor scores are materialized and why indicator loadings are kept strictly separate from structural paths.

---

## The three construct paradigms

Before writing formulas, it is critical to distinguish the causal direction
between the latent variable $\eta$ and its observed variables [@Bollen1989;
@BollenLennox1991; @GraceBollen2005]:

```{r construct-table, echo = FALSE, results = "asis"}
cat("
| Construct type | Causal topology | Interpretation | Declaration |
| :--- | :--- | :--- | :--- |
| **Formative / Composite** | Indicators $\\to$ Construct | Indicators *define* or *cause* the construct (e.g. Diet + Activity $\\to$ Body Condition). | `drm_composite()` or `drm_latent(type = \"formative\")` |
| **Reflective** | Latent $\\to$ Indicators | Indicators are *manifestations* of a common underlying factor (e.g. Boldness $\\to$ Latency, Exploration, Aggression). | `drm_latent(type = \"reflective\")` |
| **MIMIC** | Causes $\\to$ Latent $\\to$ Indicators | Multiple causes drive an unobserved state, which is reflected in multiple observable proxies (Jöreskog & Goldberger 1975). | `drm_latent(type = \"mimic\", causes = ...)` |
")
```

---

## 1. Formative constructs: `drm_composite()`

A formative construct is a deterministic or weighted combination of observed
causes [@BollenLennox1991; @Grace2008]. Because the indicators cause the
construct, indicators are not required to correlate positively with each other,
and internal-consistency reliability is not a validity requirement.

```{r formative-example}
# Simulated morphometric data for 250 birds
n <- 250
dat <- data.frame(
  bill_length = rnorm(n, mean = 20, sd = 2),
  tarsus_length = rnorm(n, mean = 35, sd = 3),
  body_mass = rnorm(n, mean = 15, sd = 1.5),
  temp = rnorm(n, mean = 18, sd = 4)
)

# Build an equal-weighted composite index of body size
size_comp <- drm_composite(
  name = "structural_size",
  indicators = c("bill_length", "tarsus_length", "body_mass"),
  method = "fixed",
  data = dat,
  standardize = TRUE
)
size_comp
```

When passed to `drm_sem()`, the composite score is materialized into the data
frame prior to node fitting. Structural nodes can use `structural_size` as an
ordinary predictor or response.

---

## 2. Reflective constructs: `drm_latent(type = "reflective")`

In a reflective construct, an unobserved latent variable $\eta$ drives multiple
observable indicators:

$$y_i = \nu_i + \lambda_i \eta + \varepsilon_i, \quad \varepsilon_i \sim \mathcal{N}(0, \theta_i)$$

where $\lambda_i$ is the factor loading for indicator $i$, and $\theta_i$ is the
unique (error) variance [@Bollen1989].

### Identification constraints

Because $\eta$ is unobserved, its scale and origin are indeterminate without an
identification constraint. `drm_latent()` provides two standard identification
options:

1. **Marker variable identification (`identification = "marker"`):**
   Fixes the loading of a reference indicator to $\lambda_1 = 1$. The scale of
   $\eta$ is identical to the scale of that marker indicator.
2. **Unit-variance identification (`identification = "unit_variance"`):**
   Constrains the latent variance to $\text{Var}(\eta) = 1$. All indicator loadings
   $\lambda_i$ are estimated freely as factor loadings.

```{r reflective-example}
# Generate data driven by a single latent trait: 'boldness'
boldness_true <- rnorm(n, mean = 0, sd = 1.5)
dat$exploratory_speed <- 1.0 * boldness_true + rnorm(n, sd = 0.5)
dat$novel_object_time <- 0.8 * boldness_true + rnorm(n, sd = 0.6)
dat$trap_latency      <- 0.6 * boldness_true + rnorm(n, sd = 0.7)

# Declare a reflective latent with marker identification:
refl_marker <- drm_latent(
  name = "boldness",
  indicators = c("exploratory_speed", "novel_object_time", "trap_latency"),
  type = "reflective",
  identification = "marker",
  marker = "exploratory_speed",
  data = dat
)
summary(refl_marker)
```

Notice that `exploratory_speed` has a loading of exactly $1.0$, fixing the
metric of the latent construct.

Alternatively, using `identification = "unit_variance"`:

```{r reflective-unit-var}
refl_uvar <- drm_latent(
  name = "boldness_std",
  indicators = list(
    drm_indicator("exploratory_speed"),
    drm_indicator("novel_object_time"),
    drm_indicator("trap_latency")
  ),
  type = "reflective",
  identification = "unit_variance",
  data = dat
)
summary(refl_uvar)
```

---

## 3. MIMIC models: Multiple Indicators, Multiple Causes

A **MIMIC** (Multiple Indicators, Multiple Causes) construct [@JoreskogGoldberger1975]
combines structural causes and reflective indicators in a single unified block:

$$\eta = \sum_{j} \gamma_j x_j + \zeta, \quad \zeta \sim \mathcal{N}(0, \psi)$$
$$y_i = \lambda_i \eta + \varepsilon_i, \quad \varepsilon_i \sim \mathcal{N}(0, \theta_i)$$

Environmental causes $x_1, x_2, \dots$ predict the latent variable $\eta$, which
in turn manifests in observed proxy indicators $y_1, y_2, \dots$.

```{r mimic-example}
# Environmental causes of condition
dat$food_availability <- rnorm(n)
dat$parasite_load     <- rnorm(n)
condition_true <- 0.7 * dat$food_availability - 0.5 * dat$parasite_load + rnorm(n, sd = 0.4)

# Reflective physiological indicators of condition
dat$hematocrit <- 1.0 * condition_true + rnorm(n, sd = 0.3)
dat$glucose    <- 0.75 * condition_true + rnorm(n, sd = 0.35)
dat$plasma_fat <- 0.55 * condition_true + rnorm(n, sd = 0.4)

# Downstream response: clutch size with location-scale effects
dat$clutch_size <- rnorm(
  n,
  mean = 4.0 + 0.8 * condition_true,
  sd = exp(0.1 - 0.4 * condition_true)
)

# Declare MIMIC construct:
mimic_cond <- drm_latent(
  name = "condition",
  causes = c("food_availability", "parasite_load"),
  indicators = list(
    drm_indicator("hematocrit", marker = TRUE),
    drm_indicator("glucose"),
    drm_indicator("plasma_fat")
  ),
  type = "mimic",
  identification = "marker",
  data = dat
)
summary(mimic_cond)
```

---

## 4. Construct reliability: $\alpha$ versus Raykov's $\rho$

Assessing construct quality is essential before interpreting structural
pathways. `drmSEM` provides both classical Cronbach's $\alpha$ and modern
composite reliability:

### Cronbach's $\alpha$

$$\alpha = \frac{k}{k - 1} \left(1 - \frac{\sum_{i=1}^k \sigma_i^2}{\sum_{i=1}^k \sum_{j=1}^k \sigma_{ij}}\right)$$

Cronbach's $\alpha$ [@Cronbach1951] assumes **essential $\tau$-equivalence**
(all indicators share identical factor loadings $\lambda_1 = \lambda_2 = \dots = \lambda_k$).
When loadings differ (as is almost universally true in real biological data),
$\alpha$ **underestimates** scale reliability [@Raykov1997].

### Raykov's composite reliability $\rho$ (McDonald's $\omega$)

$$\rho = \frac{\left(\sum_{i=1}^k \lambda_i\right)^2 \text{Var}(\eta)}{\left(\sum_{i=1}^k \lambda_i\right)^2 \text{Var}(\eta) + \sum_{i=1}^k \theta_i}$$

Raykov's $\rho$ [@Raykov1997; @McDonald1999] allows congeneric indicators with
unequal loadings $\lambda_i$ and unique error variances $\theta_i$.

```{r reliability-demo}
# Compute standalone metrics on an indicator matrix
ind_df <- dat[c("hematocrit", "glucose", "plasma_fat")]

alpha_val <- drm_cronbach_alpha(ind_df)
rho_val   <- drm_raykov_rho(ind_df)

c("Cronbach's alpha" = alpha_val, "Raykov's rho" = rho_val)
```

When indicator loadings vary substantially, Raykov's $\rho$ gives the honest,
unattenuated measure of composite reliability.

---

## 5. Assembling the SEM and estimating distributional pathways

In `drmSEM`, latent and composite constructs integrate seamlessly into
piecewise structural equation models via the `latents =` or `composites =`
arguments:

```{r fit-sem, eval = has_engine}
sem <- drm_sem(
  # Node 1: Condition driven by environmental causes
  condition = drm_node(
    drmTMB::bf(condition ~ food_availability + parasite_load),
    family = stats::gaussian()
  ),
  # Node 2: Clutch size has location (mu) AND scale (sigma) predicted by condition
  clutch_size = drm_node(
    drmTMB::bf(
      clutch_size ~ condition,
      sigma ~ condition
    ),
    family = stats::gaussian()
  ),
  data = dat,
  latents = mimic_cond
)

summary(sem)
```

### Inspecting structural paths vs indicator loadings

`drmSEM` strictly separates structural paths between nodes from indicator
measurement loadings:

```{r paths-and-loadings, eval = has_engine}
# Structural regression coefficients between nodes
paths(sem)

# Indicator loadings (measurement model)
loadings(sem)

# Construct reliability summary across all latent blocks
reliability(sem)
```

---

## 6. Distributional effect propagation through latent nodes

A central strength of `drmSEM` is that causal effects are **component-labelled**.
A path from a latent construct to a downstream node can target the **mean**
($\mu$), the **dispersion / scale** ($\sigma$), the **shape** ($\nu$), or the
**zero-inflation** ($\text{zi}$) component.

Suppose we intervene on `food_availability`, shifting it by $+1.0$ standard
deviation. What happens downstream to `clutch_size`?

```{r effects-sim, eval = has_engine}
# Simulate total interventional effects of food_availability on clutch_size
eff <- total_effects(sem, from = "food_availability", to = "clutch_size", nsim = 200)
eff
```

### Interpreting component-labelled pathways

- **Path `condition -> clutch_size (mu)`**: Increases the expected mean clutch size.
- **Path `condition -> clutch_size (sigma)`**: Modulates the residual variation or reproductive risk among individuals. **This is not a mean effect.** A negative $\sigma$ path indicates that birds in higher condition have more predictable, less erratic clutch sizes.

By simulating forward through the structural graph, `drmSEM` correctly
propagates interventions through the latent construct to all modelled distributional
parameters simultaneously.

---

## Summary of best practices

1. **Choose the right construct model**: Use `drm_composite()` when indicators are causes/inputs (formative); use `drm_latent()` when indicators are reflections/proxies of an underlying factor.
2. **Select an identification rule**: Use `marker` when you want the latent construct to inherit the scale of a primary reference trait; use `unit_variance` when you want a standardized factor score.
3. **Check composite reliability**: Report Raykov's $\rho$ alongside Cronbach's $\alpha$ to ensure proxy indicators measure a congeneric construct reliably ($\rho \ge 0.70$).
4. **Distinguish structural paths from loadings**: In publications, report measurement loadings ($\lambda_i$) separately from structural regressions ($\beta_{j}$).
5. **Embrace distributional targets**: Use location-scale formulas (`sigma ~ eta`) when latent quality stabilizes or destabilizes downstream biological traits.

---

## References
