---
title: "Graph-derived piecewise imputation for missing data (drm_sem)"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Graph-derived piecewise imputation for missing data (drm_sem)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Model-fitting chunks use the drmTMB engine and are eval-gated; the pure-R
# declaration grammar and diagnostic tables run.
has_engine <- requireNamespace("drmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
library(drmSEM)
set.seed(20260828)
```

Missing data is pervasive in ecology, evolutionary biology, and the environmental
sciences [@LittleRubin2019; @Enders2010]. In mediation chains and structural equation
models ($X \to M \to Y$), intermediate traits $M$ (such as hormone titers, metabolic
rates, behavioural assays, or tissue samples) are often missing on a substantial
fraction of individuals.

When mediator $M$ is missing conditionally on the outcome $Y$ (**Missing at Random**, MAR),
dropping incomplete rows via **complete-case analysis induces severe bias** in both
the direct and indirect paths.

`drmSEM` provides **graph-derived piecewise imputation** via `impute = "auto"` in
`drm_sem()` and `drm_psem()`. This vignette explains:

1. Why complete-case analysis fails in causal mediation chains.
2. How `drmSEM` derives imputation models directly from the causal DAG without ad-hoc user choices.
3. Supported response and predictor families (`Gamma`, `lognormal`, `student`, `nbinom2`, `ZIP`, `gaussian`).
4. Multi-parent imputation ($k = 2$ incomplete parents) and engine constraints.
5. Inspection tools: `imputation()`, `imputed()`, and `check_sem()`.
6. The honest scientific boundary: why piecewise graph-derived imputation is not joint FIML, and how within-node uncertainty is propagated.

---

## 1. The problem: mediator missingness in causal chains

Consider a classic mediation chain where an environmental driver $X$ influences
a physiological mediator $M$, which in turn affects fitness $Y$:

$$X \to M \to Y$$

```{r mar-data}
# Simulate a mediation chain with outcome-dependent missingness (MAR)
n <- 400
x <- rnorm(n, mean = 0, sd = 1)
m_true <- 0.7 * x + rnorm(n, sd = 0.5)
y <- 0.5 * m_true + 0.4 * x + rnorm(n, sd = 0.5)

dat <- data.frame(x = x, m = m_true, y = y)

# Induce MAR missingness: individuals with low fitness (Y) are less likely to be sampled for M
p_miss <- plogis(-0.5 - 1.2 * scale(y))
dat$m[runif(n) < p_miss] <- NA

# Check missingness proportion
mean(is.na(dat$m))
```

If we fit this system with standard complete-case analysis (`impute = "none"`),
any row with missing $M$ is dropped from the regression of $Y$ on $M$ and $X$.
Because missingness depends on $Y$, complete-case estimates of $M \to Y$ and
$X \to Y$ are substantially biased.

---

## 2. Graph-derived automatic imputation (`impute = "auto"`)

In traditional structural equation modelling, handling missing predictors
requires either:
- Specifying an ad-hoc imputation model (risking uncongeniality with the analysis model), or
- Running external chained equations (MICE) with manual variable selections.

In `drmSEM`, **the causal DAG already specifies the imputation model**.
Because $M$ is an endogenous node in the SEM with its own formula and family
($M \sim X$, $\text{family} = \dots$), `drmSEM` automatically borrows node $M$'s
model and wires it directly into node $Y$'s likelihood as a modelled missing
predictor (`mi(m)`):

```{r auto-impute, eval = has_engine}
sem_imputed <- drm_sem(
  m = drm_node(drmTMB::bf(m ~ x), family = stats::gaussian()),
  y = drm_node(drmTMB::bf(y ~ m + x), family = stats::gaussian()),
  data = dat,
  impute = "auto"
)

summary(sem_imputed)
```

No manual imputation formulas or custom missing-data scripts are required:
the DAG supplies the exact parent set and distributional family for $M$.

---

## 3. Supported continuous and discrete families

`drmSEM` supports graph-derived imputation across a wide range of response
and predictor distributions:

### Continuous families
- **Gaussian (`gaussian`)**: Standard continuous location-scale responses.
- **Gamma (`Gamma` / `gamma`)**: Positive continuous traits (e.g. body size, metabolic rate, dispersal distance).
- **Log-normal (`lognormal`)**: Skewed positive measurements.
- **Student-$t$ (`student`)**: Heavy-tailed robust continuous outcomes.
- **Beta (`beta`)**: Proportions and bounded indices on $(0, 1)$.

### Discrete and count families
- **Poisson (`poisson`)**: Unbounded count data (e.g. egg counts, territory visits).
- **Negative Binomial (`nbinom2`)**: Overdispersed ecological count responses with quadratic variance $\mu + \mu^2/\theta$.
- **Zero-Inflated Poisson (`ZIP` / `zi_poisson`)**: Count data with excess structural zeros (e.g. parasite counts, mating successes).
- **Binomial & Beta-Binomial (`binomial`, `beta_binomial`)**: Binary trials and overdispersed success proportions.

```{r count-example, eval = has_engine}
# Example with a Gaussian mediator and Negative Binomial count response
dat_count <- data.frame(
  temp = rnorm(n),
  x_env = rnorm(n)
)
dat_count$biomass <- 0.6 * dat_count$temp + rnorm(n, sd = 0.5)
dat_count$parasites <- rpois(n, lambda = exp(0.5 + 0.4 * dat_count$biomass - 0.3 * dat_count$x_env))

# Introduce missingness in biomass
dat_count$biomass[runif(n) < 0.25] <- NA

sem_count <- drm_sem(
  biomass   = drm_node(drmTMB::bf(biomass ~ temp), family = stats::gaussian()),
  parasites = drm_node(drmTMB::bf(parasites ~ biomass + x_env), family = drmTMB::nbinom2()),
  data = dat_count,
  impute = "auto"
)

paths(sem_count)
```

---

## 4. Multi-parent imputation ($k = 2$) and engine constraints

When a downstream node $Y$ has **two incomplete endogenous parents**
($X \to M_1 \to Y$ and $X \to M_2 \to Y$), `drmSEM` automatically derives
independent modelled imputation terms for both parents:

```{r two-parent, eval = has_engine}
# Two incomplete Gaussian parents
dat2 <- data.frame(x = rnorm(n))
dat2$m1 <- 0.6 * dat2$x + rnorm(n, sd = 0.5)
dat2$m2 <- -0.4 * dat2$x + rnorm(n, sd = 0.5)
dat2$y  <- 0.5 * dat2$m1 + 0.3 * dat2$m2 + 0.2 * dat2$x + rnorm(n, sd = 0.5)

# Missingness in both parents
dat2$m1[runif(n) < 0.20] <- NA
dat2$m2[runif(n) < 0.20] <- NA

sem_2p <- drm_sem(
  m1 = drm_node(drmTMB::bf(m1 ~ x), family = stats::gaussian()),
  m2 = drm_node(drmTMB::bf(m2 ~ x), family = stats::gaussian()),
  y  = drm_node(drmTMB::bf(y ~ m1 + m2 + x), family = stats::gaussian()),
  data = dat2,
  impute = "auto"
)

paths(sem_2p)
```

### Engine constraints and clear error reporting

`drmSEM` enforces strict honesty regarding engine limits:
- **$k > 2$ incomplete parents**: The fitting engine currently limits simultaneous modelled imputation to $k \le 2$ terms per node. Attempting $k > 2$ aborts with an actionable error directing the user to complete or impute upstream predictors.
- **$k = 2$ on non-Gaussian responses**: Multi-parent imputation currently requires a Gaussian response node; non-Gaussian $k = 2$ cells fail loudly with clear diagnostics.

---

## 5. Inspecting imputation models and diagnostics

`drmSEM` provides three dedicated diagnostic accessors:

### 1. `imputation(sem)`
Summarizes the exact graph-derived imputation plan for each node:

```{r inspect-plan, eval = has_engine}
imputation(sem_imputed)
```

### 2. `imputed(sem)`
Extracts observation-level imputed predictions, standard errors, and status flags:

```{r inspect-imputed, eval = has_engine}
imp_vals <- imputed(sem_imputed)
head(imp_vals)
```

Observations that were originally missing carry `observed = FALSE` and report
imputed values along with their model-based standard errors.

### 3. `check_sem(sem)`
Performs a comprehensive health check across the entire SEM, reporting
sample sizes, convergence status, covariance availability, and missingness:

```{r inspect-check, eval = has_engine}
check_sem(sem_imputed)
```

---

## 6. Honest piecewise bounds versus FIML

It is critical to distinguish **piecewise graph-derived imputation** from
**Full-Information Maximum Likelihood (FIML)**:

1. **Within-node parameter estimation**:
   In piecewise SEM, node $Y$ re-estimates the missing parent model ($M \sim X$)
   *within its own likelihood function*. It does not rigidly fix the parameters
   to node $M$'s point estimates.
2. **Uncertainty propagation**:
   Imputation uncertainty for $M$ is fully accounted for within node $Y$'s
   joint Hessian (standard errors for $M \to Y$ and $X \to Y$ correctly reflect
   missing-data uncertainty).
3. **No across-node joint likelihood**:
   Because `drmSEM` is piecewise, it does not fit a single global joint covariance
   matrix across all endogenous nodes. It must never be described as whole-SEM FIML.

This approach provides a robust, principled solution for missing mediators that
eliminates MAR attrition bias while respecting the piecewise architecture.

---

## Summary of best practices

1. **Activate auto-imputation**: Use `drm_sem(..., impute = "auto")` whenever endogenous mediators contain missing observations.
2. **Check the imputation plan**: Run `imputation(sem)` to verify that derived parent formulas match your theoretical DAG.
3. **Inspect individual imputations**: Use `imputed(sem)` to verify that imputed values lie within reasonable biological bounds.
4. **Audit model health**: Run `check_sem(sem)` to confirm node convergence and Hessian positive-definiteness across all sub-models.

---

## References
