---
title: "Model selection with CBIC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model selection with CBIC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# This article explains the workflow without fitting models during R CMD check
# or pkgdown. The validation table is read from the cached live-lane study.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)

source_results_path <- file.path("inst", "validation",
                                 "validation-results.rds")
installed_results_path <- system.file("validation", "validation-results.rds",
                                      package = "drmSEM")
results_path <- if (file.exists(source_results_path)) {
  source_results_path
} else {
  installed_results_path
}
val <- if (nzchar(results_path) && file.exists(results_path)) {
  readRDS(results_path)
} else {
  NULL
}

library(drmSEM)
```

Use model selection when you have a small set of biologically plausible DAGs and
want to ask which graph is best supported after each endogenous node has been
fitted with `drmTMB`. The workflow is confirmatory: you name the candidate DAGs,
fit the same data under each candidate, and compare their d-separation evidence.
It is not an all-subsets search over every possible graph.

`drmSEM` now uses **CBIC** as the default ranking criterion. CBIC is a BIC-style
criterion built on Fisher's C,

\[
\mathrm{CBIC} = C + k\log(n),
\]

where \(C\) is Fisher's C statistic from the d-separation test, \(k\) is the
number of fitted fixed-effect paths, and \(n\) is the number of observations.
The stronger penalty matters because a harmless over-fitted DAG can have a very
small C statistic. CBIC asks a stricter question: is the extra path worth its
cost?

## Build a candidate set

Suppose the scientific question is whether temperature acts on fitness directly
or through body size. The candidate set might contain a direct model, a mediated
model, and a deliberately over-fitted model with both routes.

```{r candidate-set}
models <- drm_model_set(
  direct = drm_dag(
    size ~ temp,
    fitness ~ temp
  ),
  mediated = drm_dag(
    size ~ temp,
    fitness ~ size
  ),
  direct_plus_mediated = drm_dag(
    size ~ temp,
    fitness ~ temp + size
  )
)
```

Each `drm_dag()` is an unfitted candidate graph. The formulas describe the
structural parents of each endogenous node. A fair comparison keeps the response
set and data set fixed across candidates, then changes the arrows that represent
the competing biological explanations.

## Fit and rank the models

`compare()` fits every candidate with `drm_sem()`, runs the d-separation test,
combines the independence-claim p-values into Fisher's C, and adds both CBIC and
CICc scores.

```{r compare}
cmp <- compare(
  models,
  data = dat,
  family = list(
    size = stats::gaussian(),
    fitness = drmTMB::nbinom2()
  )
)

cmp
```

The printed table contains both criteria:

- `CBIC`, `dCBIC`, and `wCBIC` are the default ranking score, delta, and weight.
- `CICc`, `dCICc`, and `wCICc` are retained for the phylopath-style support
  ranking.
- `weight` is the weight for the selected criterion. With the default
  `compare()`, it is the same as `wCBIC`.

The top model is the row with the lowest CBIC. `best()` returns that fitted SEM.

```{r best-average}
top <- best(cmp)
paths(top)

avg <- average(cmp)
avg
```

`average()` returns criterion-weighted standardized paths. With the default
ranking, those are CBIC-weighted paths. A path that appears in only some
candidates is averaged conditionally over the models that contain it, with
`weight_sum` showing how much model weight supported that path.

## When to ask for CICc

CICc remains available:

```{r cicc}
cmp_cicc <- compare(
  models,
  data = dat,
  family = list(
    size = stats::gaussian(),
    fitness = drmTMB::nbinom2()
  ),
  criterion = "CICc"
)
```

Use CICc when you want a phylopath-style support ranking or a sensitivity check.
Do not read CICc as the default true-DAG recovery claim. In the validation study
below, CICc often retained an over-fitted rival that added a harmless direct
path, while CBIC selected the generating DAG.

## What the validation showed

The model-selection validation study generated data from a true chain
`x -> m -> y`, compared that truth against an over-fitted rival and a
missing-edge rival, and repeated the comparison 300 times at each sample size.
The current cached results are:

```{r validation-table, echo = FALSE, eval = TRUE}
if (!is.null(val) && !is.null(val$model_selection)) {
  sel_src <- val$model_selection
  if (!"criterion" %in% names(sel_src)) {
    sel_src$criterion <- "CICc"
  }
  sel_tab <- sel_src[order(sel_src$criterion, sel_src$n),
                     c("criterion", "n", "selection_rate",
                       "mean_truth_weight", "missing_edge_rate", "n_rep",
                       "acceptance")]
  knitr::kable(
    sel_tab,
    digits = 3,
    col.names = c("criterion", "n", "truth selection",
                  "mean truth weight", "missing-edge rate", "reps", "pass"),
    caption = "Model-selection recovery in the cached validation study."
  )
} else {
  cat("No cached model-selection validation table is available in this build.")
}
```

CBIC passed the recovery gate in this grid: it selected the true DAG in 92.7% of
replicates at `n = 300` and 97.3% at `n = 1000`, and it never selected the
missing-edge model. CICc also avoided the missing-edge model, but it selected
the true DAG only 61.0% and 57.3% of the time because the over-fitted rival kept
substantial support. That is why CBIC is the default.

## A practical reading rule

Read the comparison table in this order:

1. Check whether any candidate has a poor d-separation result. A model with a
   violated missing-edge claim is usually not a good scientific explanation.
2. Use CBIC and `wCBIC` to choose the parsimonious candidate.
3. Inspect CICc as a support sensitivity check. If CICc gives a harmless
   over-fitted model non-trivial support, report that as uncertainty about a
   weak extra path, not as evidence that the extra path is needed.
4. Use `average()` when the substantive paths are more important than the
   identity of a single top model.

This keeps the workflow honest: CBIC answers the "which DAG should I carry
forward?" question, while CICc remains a useful view of near-ties and
over-fitted alternatives.
