---
title: "Phylogenetic distributional SEM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phylogenetic distributional SEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# drmSEM never fits its own likelihoods: every endogenous node is one drmTMB
# fit, and drmTMB compiles TMB (C++) from source. The phylogenetic examples
# below additionally need a tree (the `ape` package). To keep this vignette
# building under R CMD check / pkgdown without either dependency installed,
# the model-fitting chunks are shown for illustration but NOT evaluated. Flip
# `has_engine` to test for both packages once integration is validated, e.g.
# requireNamespace("drmTMB", quietly = TRUE) && requireNamespace("ape", quietly = TRUE).
has_engine <- FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_engine
)
library(drmSEM)
```

## What is a *phylogenetic distributional* SEM?

**`drmSEM` is the only SEM tool that runs d-separation and a distributional
effect decomposition over a phylogenetically-structured DAG** -- it does
phylogenetic *distributional* path analysis, where shared ancestry is a
controlled-for random effect and a causal arrow can target the *spread*,
*zero-inflation*, or *shape* of a trait, not only its mean.

That sentence is best understood by contrast with the two neighbouring tools an
evolutionary ecologist already knows:

| Package     | What it does                                                       |
| ----------- | ----------------------------------------------------------------- |
| `phylopath` | mean-only **confirmatory path analysis** on a phylogeny (d-sep + C) |
| `phylosem`  | **joint** estimation of a phylogenetic comparative model *and* a SEM |
| `drmSEM`    | phylogenetic **distributional** SEM: per-node drmTMB fits, d-sep and effect decomposition over the fixed-effect DAG, with phylogeny as a structured random effect |

`phylopath` answers "which DAG of *mean* relationships survives the data?".
`phylosem` couples a Brownian-motion (or OU) process to a latent-variable SEM.
`drmSEM` keeps the piecewise, observed-variable, one-fit-per-node design of the
core package, and adds shared ancestry as just another random-effect term inside
each node's likelihood -- so the same machinery that already separates a
`sigma` path from a `mu` path now also works *after correcting for phylogeny*.

## The key idea: phylogeny is a structured random effect, not an edge

In `drmSEM` each endogenous node is exactly one `drmTMB` fit (see the
`vignette("drmSEM")` introduction). To make a node *phylogenetic*, you add one
random-effect term to its formula:

```
phylo(1 | species, tree = phy)
```

This tells the fitting engine to impose a phylogenetic covariance (derived from
the tree `phy`) on a species-level random intercept, exactly as a phylogenetic
mixed model would. Crucially, **`drmSEM` treats this term as a random effect,
not a causal edge.** Its formula parser strips structured-effect markers --
`phylo`, `spatial`, `animal`, `gr`, and friends -- before turning the remaining
predictors into typed edges (`R/utils.R`, `drm_strip_markers()`). So adding
`phylo(1 | species, tree = phy)` to a node:

- does **not** add `species` (or the tree) as a parent in the DAG;
- does **not** appear in `paths()` as a structural coefficient;
- **is** carried inside the node's likelihood, so every fixed-effect coefficient
  is estimated *conditional on* shared ancestry.

The practical consequence is the one you want: `paths()`, `basis_set()`,
`dsep()`, `fisher_c()`, and the effect decomposition all operate on the
**fixed-effect DAG**, while the phylogenetic signal is soaked up by the random
effect underneath. You get phylogenetically-corrected path coefficients and
phylogenetically-corrected independence tests "for free", because the correction
lives where it belongs -- in each node's fit -- rather than in the graph.

## A worked system: `size -> abundance -> fitness`, on a tree

We reuse the chain motif from the core vignette, but now every node is measured
across **species** related by a phylogeny `phy`, and every node carries a
phylogenetic random intercept. Conceptually:

- **`size`** depends on temperature, with temperature also acting on the
  *scale* of size (`sigma(size) ~ temp`) -- a distributional path -- plus a
  phylogenetic intercept.
- **`abundance`** is a count depending on `size` and `temp`, modelled as a
  negative binomial with habitat acting on *zero-inflation*
  (`zi(abundance) ~ habitat`), again with a phylogenetic intercept.
- **`fitness`** is a count (e.g. lifetime offspring) depending on `abundance`
  and `size`, with its own phylogenetic intercept.

Temperature is exogenous and reaches `fitness` only indirectly -- through the
*mean*, the *scale*, and the *zero-inflation* of the intermediate nodes -- and
the whole system is evaluated after correcting for shared ancestry.

### The tree and the data

In a real analysis you would read a tree (for example with `ape::read.tree()`)
and align it to a species-level data frame. We sketch that here; this chunk is
not evaluated because it needs `ape`. Two requirements: the tree must be
**ultrametric** (equal root-to-tip distances; rescale a non-ultrametric tree
with e.g. `ape::compute.brlen(tree, "Grafen")`), and every level of `species`
must be a tip label of the tree.

```{r tree, eval = has_engine}
library(ape)
# A phylogeny over the species in `dat`; tip labels must match dat$species.
phy <- read.tree("my_tree.nwk")

# `dat` has one row per observation, with a `species` column whose levels are
# tip labels of `phy`, plus the measured traits and covariates.
str(dat)
#> 'data.frame': ... obs. of  variables:
#>  $ species  : chr  "sp1" "sp2" ...
#>  $ temp     : num ...
#>  $ habitat  : Factor ...
#>  $ size     : num ...
#>  $ abundance: int ...
#>  $ fitness  : int ...
```

### Build the phylogenetic SEM

The only change relative to a non-phylogenetic SEM is the
`phylo(1 | species, tree = phy)` term added to each node's formula. Everything
else -- families, distributional sub-formulas, the chain structure -- is
ordinary `drmSEM`.

```{r build}
sem <- drm_sem(
  size = drm_node(
    drmTMB::bf(
      size ~ temp + phylo(1 | species, tree = phy),
      sigma ~ temp                                  # distributional path: temp -> sigma(size)
    ),
    family = stats::gaussian()
  ),
  abundance = drm_node(
    drmTMB::bf(
      abundance ~ size + temp + phylo(1 | species, tree = phy),
      zi ~ habitat                                  # distributional path: habitat -> zi(abundance)
    ),
    family = drmTMB::nbinom2()
  ),
  fitness = drm_node(
    drmTMB::bf(
      fitness ~ abundance + size + phylo(1 | species, tree = phy)
    ),
    family = drmTMB::nbinom2()
  ),
  data = dat
)
```

### The DAG is the *fixed-effect* DAG

Because the `phylo(...)` term is stripped before edge extraction, `paths()`
shows only the structural, fixed-effect coefficients -- one row per modelled
component. The phylogenetic intercepts are present in every node's fit but
absent from the path table, which is exactly what you want for causal reasoning.

```{r paths}
paths(sem)
#> Note: no `species` or tree rows appear here. Each `temp` path into `size`
#> still appears twice -- once on `mu`, once on `sigma` -- and the
#> `habitat -> zi(abundance)` path is labelled as a zero-inflation path, all
#> estimated conditional on shared ancestry.
```

`plot()` draws the same fixed-effect DAG with component-labelled edges. Phylogeny
is *not* drawn as a node or an arrow; it is a property of the fits.

```{r plot, fig.alt = "Directed acyclic graph of the size, abundance and fitness nodes with component-labelled edges; phylogeny is a per-node random effect and is not drawn as an edge."}
plot(sem)
```

### Phylogenetic d-separation

The independence claims are tested exactly as in the non-phylogenetic case --
the **any-component LRT** -- but each refit now also carries the phylogenetic
random effect, so the tests are phylogenetically corrected. A missing arrow
`X -> Y` asserts that `X` has no effect on *any* modelled component of `Y` *after*
accounting for shared ancestry.

```{r dsep}
dsep(sem)
```

```{r fisherc}
fisher_c(sem)
```

A small Fisher's C p-value flags a missing path -- including a missing path into
a *non-mean* component -- that survives phylogenetic correction. Neither a
mean-only phylogenetic tool nor a non-phylogenetic distributional tool could
raise that flag.

### Phylogenetic *distributional* effects

This is the novel surface. `indirect_effects()` propagates a change in `temp`
through the fitted DAG by Monte-Carlo simulation, so effects flowing through a
mediator's `sigma`, `zi`, or shape are included -- and because every node's
draws come from a phylogenetically-corrected fit, the decomposition is a
**phylogenetic distributional** decomposition.

```{r indirect}
indirect_effects(sem, from = "temp", to = "fitness")
#> total_path             -- full simulated temp -> fitness effect
#> direct                 -- controlled direct effect
#> indirect               -- total_path - direct
#> mean_mediated          -- part carried by mediator means
#> distribution_mediated  -- EXTRA effect via sigma(size) and zi(abundance),
#>                           i.e. the phylogenetic distributional channel
```

A non-trivial `distribution_mediated` row is the signature of this framework on
a tree: after correcting for phylogeny, temperature still reaches fitness partly
because it changes the *spread* of size and (through habitat structure) the
*zeros* of abundance -- not only their means. `phylopath` cannot see that
channel because it is mean-only; a non-phylogenetic distributional SEM would
confound it with ancestry.

## What works now vs roadmap

`drmSEM`'s phylogenetic support is staged. See
`docs/design/06-phylogenetic-sem.md` for the full design and the honest
non-goals.

**Phase 1 -- phylogenetic nodes (works today).** A node can carry a
`phylo(1 | species, tree = phy)` random effect. The marker is stripped from the
causal edge set, so `paths()`, `basis_set()`, `dsep()`, `fisher_c()`, and the
effect decomposition all run over the fixed-effect DAG while each node is fitted
conditional on shared ancestry. This is the worked example above (subject to the
same drmTMB-integration validation caveat as the rest of the package).

**Phase 2 -- phylopath-style model comparison (roadmap).** Ranking a *set* of
candidate phylogenetic DAGs by a C-statistic information criterion (the
`phylopath` workflow), with model-averaged paths, is planned but not yet
implemented.

**Phase 3 -- evolutionary process models (roadmap).** First-class choice of the
phylogenetic covariance -- Brownian motion (BM), Ornstein-Uhlenbeck (OU), and
Pagel's lambda / kappa transformations -- as a per-node option, rather than the
single default correlation structure available now.

**Phase 4 -- distributional phylo paths (roadmap).** Letting the phylogenetic
*structure itself* differ by component (e.g. a distinct evolutionary process on
`sigma` or `zi`), so that "how variance evolves" becomes a modelled,
component-labelled target.

Per `docs/design/06-phylogenetic-sem.md`, the non-goals are deliberate and
inherited from the charter: `drmSEM` will not fit its own phylogenetic
likelihoods (drmTMB remains the only engine), will not jointly estimate a tree,
and will not replace `phylosem` for joint PCM+SEM estimation. The aim is narrow
and complementary -- phylogenetic *distributional* path analysis -- not a
re-implementation of the comparative-methods stack.

## Recap

- Phylogeny enters as a **structured random effect**
  `phylo(1 | species, tree = phy)`, one per node -- not as a DAG edge.
- `drmSEM` **strips** that term from the causal edge set, so `paths()`,
  `dsep()`, `fisher_c()`, and the effect decomposition operate on the
  **fixed-effect DAG**, phylogenetically corrected.
- The novel surface is **phylogenetic distributional** paths: effects into
  `sigma`, `zi`, or shape that survive correction for shared ancestry.
- Phase 1 (phylo nodes) works today; Phases 2-4 (phylopath-style comparison,
  BM/OU/lambda/kappa, distributional phylo paths) are roadmap. See
  `docs/design/06-phylogenetic-sem.md`.
</content>
</invoke>
