---
title: "Effect-interval coverage and model-selection recovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Effect-interval coverage and model-selection recovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# This is a PRECOMPUTED vignette. Every result shown below is read from a
# cached study object built in the live-drmTMB lane (see the "How this was
# generated" section). The global eval = FALSE guarantees that R CMD check,
# pkgdown, and any reader rebuilding the vignette NEVER fit a model here.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)

source_results_path <- file.path("inst", "validation",
                                 "validation-results.rds")
installed_results_path <- system.file("validation", "validation-results.rds",
                                      package = "drmSEM")
results_path <- if (file.exists(source_results_path)) {
  source_results_path
} else {
  installed_results_path
}
val <- if (nzchar(results_path) && file.exists(results_path)) {
  readRDS(results_path)
} else {
  NULL
}
```

This page asks two practical validation questions: do simulated effect intervals
cover known effects at the advertised rate, and does model comparison recover the
true DAG from a small candidate set? All results below are **precomputed on live
drmTMB; this document performs no model fitting.** The study object is
regenerated only in the drmTMB lane and ships as a cached `.rds` inside the
package.

```{r guard, eval = TRUE, echo = FALSE, results = "asis"}
# Graceful degrade: render an explanatory note (not an error) and stop early if
# the cache has not been built yet, so the doc still renders in the
# no-engine lane.
if (is.null(val)) {
  cat(
    "> **The effect-interval and model-selection validation study has not yet been run in the live-drmTMB lane.**\n",
    ">\n",
    "> This page is a precomputed scaffold. Once",
    "`inst/validation/generate.R`",
    "is run with the drmTMB engine, the cached results will appear here:",
    "the effect-CI coverage table and the model-selection recovery table.",
    "The study design is shown below for transparency.\n",
    sep = " "
  )
  knitr::knit_exit()
}
```

## What is being tested

The validation matrix (`docs/design/11-validation-matrix.md`) established
**point recovery**: the machinery returns the right *number* on a known-answer
DGP. This page (`docs/design/12-coverage-calibration.md`) establishes the
**operating characteristics** -- does the *uncertainty* have the advertised
behaviour, and do the *inferential* procedures calibrate? These are
many-replicate studies, so the full grid runs in the live lane and is cached,
like the d-separation calibration page.

Two studies are tabulated here:

- **Effect-interval coverage.** Does `uncertainty = "parametric"` produce
  intervals that cover the true effect at the nominal rate?
- **Model-selection recovery.** Does [`best()`] select the true DAG, and how much
  Akaike weight lands on the truth?

Two related checks live elsewhere and are **not** reimplemented here:
d-separation Type-I / power is the calibration study on its own page, and
sampler dispersion vs `drmTMB::simulate()` is the standalone probe
`inst/validation/sampler-dispersion-probe.R`.

## The study design (shown, never run)

```{r design-grid, echo = TRUE, eval = FALSE}
SEED <- 20260607
GRID <- list(
  R      = 300,                 # replicate datasets per cell
  n_grid = c(300, 1000),        # sample sizes
  B      = 400,                 # parametric-bootstrap draws per fit
  level  = 0.95                 # nominal CI level
)
```

### Effect-interval coverage

A linear identity-link Gaussian chain `x -> m -> y` with a direct `x -> y` edge,
whose **true total / direct / indirect effects are known in closed form** --
products of the data-generating coefficients times the contrast width `sd(x)`:

```{r design-c1, echo = TRUE, eval = FALSE}
# data-generating coefficients
b_xm <- 0.6; b_my <- 0.5; b_xy <- 0.3
m <- rnorm(n, b_xm * x, 1)
y <- rnorm(n, b_xy * x + b_my * m, 1)

# known truth (response units of y, per sd(x))
indirect <- b_xm * b_my * sd(x)
direct   <- b_xy        * sd(x)
total    <- (b_xy + b_xm * b_my) * sd(x)

sem <- drm_sem(
  m = drm_node(drmTMB::bf(m ~ x),     family = gaussian()),
  y = drm_node(drmTMB::bf(y ~ x + m), family = gaussian()),
  data = dat
)
tot <- total_effects(sem, "x", "y", method = "simulate",
                     uncertainty = "parametric", B = 400)
dec <- indirect_effects(sem, "x", "y", uncertainty = "parametric", B = 400)
# record whether conf.low/conf.high bracket the known truth, per quantity
```

### Model-selection recovery

A true DAG (`x -> m -> y`, with the direct `x -> y` edge omitted) versus a
candidate set that includes the truth, an over-fitted rival, and a
missing-edge rival. Over replicates we record how often [`best()`] returns the
truth and the mean Akaike weight on it.

```{r design-c3, echo = TRUE, eval = FALSE}
models <- drm_model_set(
  truth        = drm_dag(m ~ x, y ~ m),       # data-generating DAG
  overfit      = drm_dag(m ~ x, y ~ x + m),   # spurious x -> y edge
  missing_edge = drm_dag(m ~ x, y ~ x)        # true m -> y arrow left out
)
cmp <- compare(models, data = dat,
               family = list(m = gaussian(), y = gaussian()))
selected <- cmp$model[which.min(cmp$CICc)]   # best() picks the lowest CICc
```

This study code is shown **for transparency only** and is never evaluated when
the vignette is built.

## Acceptance criteria

The cached study object stores these checks in `val$acceptance`:

1. Every `(n, quantity)` empirical coverage lies inside nominal +/- 2*SE.
2. Every `n` has truth selection rate >= 0.80 and missing-edge selection rate
   <= 0.05.
3. The truth selection rate is nondecreasing in `n` (within 0.05 Monte-Carlo
   jitter).

```{r acceptance-table, echo = FALSE, eval = !is.null(val)}
criteria <- val$acceptance$criteria
criteria$criterion <- sub("^C[0-9]+\\s+", "", criteria$criterion)
knitr::kable(
  criteria,
  col.names = c("Criterion", "Rule", "Pass"),
  caption = "Validation acceptance checks."
)
```

## Effect-interval coverage

Empirical coverage of the nominal `r if (!is.null(val)) val$provenance$level`
parametric interval per quantity and sample size, with the Monte-Carlo
acceptance band (nominal +/- 2*SE) and mean interval width as a secondary
efficiency diagnostic.

```{r coverage-table, echo = FALSE, eval = !is.null(val)}
cov_tab <- val$coverage[order(val$coverage$quantity, val$coverage$n),
                        c("quantity", "n", "coverage", "mc_lower", "mc_upper",
                          "mean_width", "n_rep", "acceptance")]
knitr::kable(
  cov_tab,
  digits  = 3,
  col.names = c("quantity", "n", "coverage", "MC lower", "MC upper",
                "mean width", "reps", "pass"),
  caption = sprintf(
    "Empirical coverage of the nominal %.2f parametric interval, with the +/- 2*SE Monte-Carlo band.",
    val$provenance$level
  )
)
```

## Model-selection recovery

Selection rate of the true DAG and mean Akaike weight on the truth per sample
size, with the rate at which the under-fitted (missing-edge) rival is wrongly
selected.

```{r selection-table, echo = FALSE, eval = !is.null(val)}
sel_tab <- val$model_selection[order(val$model_selection$n),
                               c("n", "selection_rate", "mean_truth_weight",
                                 "missing_edge_rate", "n_rep", "acceptance")]
knitr::kable(
  sel_tab,
  digits  = 3,
  col.names = c("n", "selection rate", "mean truth weight",
                "missing-edge rate", "reps", "pass"),
  caption = "Truth selection rate and mean Akaike weight on the truth."
)
```

## How this was generated

The cache backing this page is built by **`inst/validation/generate.R`**, run
in the live-drmTMB lane:

```sh
Rscript inst/validation/generate.R
```

That script guards on `requireNamespace("drmTMB")`, sweeps the grids above,
collects only data (coverage flags, widths, selection labels, Akaike weights --
never fitted objects), and writes a compact summary list to
`inst/validation/validation-results.rds`. The cache carries a `provenance` stamp
recording the **drmTMB version, date, R version, git SHA, replicate count, seed,
and n-grid** so every table on this page is traceable to the exact engine that
produced it.

```{r provenance, echo = FALSE, eval = !is.null(val)}
p <- val$provenance
cat(sprintf(
  "Cache provenance:\n  generated:      %s\n  drmTMB version: %s\n  drmTMB SHA:     %s\n  drmSEM version: %s\n  R version:      %s\n  git SHA:        %s\n  replicates:     %s\n  seed:           %s\n  n grid:         %s\n",
  p$generated_at, p$drmTMB_version,
  if (is.null(p$drmTMB_remote_sha)) NA_character_ else p$drmTMB_remote_sha,
  p$drmSEM_version, p$R_version, p$git_sha, p$R, p$seed,
  paste(p$n_grid, collapse = ", ")
))
```
